---
title: "CLIMER data analysis"
output: html_notebook
---

```{r setup, include=FALSE, results = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
*Installation of packages* 
```{r}
#install.packages(c("ggplot2", "Rmisc", "dplyr" "scales", "pastecs"))
Packages <- c("ggplot2", "Rmisc", "dplyr", "scales", "pastecs")
lapply(Packages, library, character.only = TRUE)
```

### Load Data
```{r Load file}
d <- read.table("Copy of Climer_Biodeg_3.txt", sep = "\t", header = TRUE, fill=TRUE)
```

```{r}
Bulk <- subset(d, Fraction = Bulk)

DOC <- select(Bulk, Date, Lake, DOC)
summary(DOC)
tapply(DOC$DOC, DOC$Lake, summary) 

```


## Data preparation
- Remove data from Stevie due to long storage time (< 10 months)
- ordering factor variables
```{r Data preparation}
d$Date <- as.Date(d$Date,format = "%d.%m.%Y")
d<-d[!(d$Student=="Stevie"),]

#d<-d[!(d$Fraction=="Bulk"),]
#d<-d[(d$Replica=="1"),] #Why only look at this? someone without replica
#d<-d[(d$Fraction=="HMW"),] # look at LMW data with high variation
#d <- subset(d, Fraction =="LMW" & Lake == "Inlet")

d$Lake <- factor(d$Lake,
                 levels = c("Inlet","Hypolimnion", "Outlet"),ordered = TRUE)
d$Fraction <- factor(d$Fraction,
                     levels = c("LMW","HMW", "Bulk"),ordered = TRUE)
d$Month <- factor(d$Month,
                  levels = c("September_17","April_17", "June_18"),ordered = TRUE)
d$Month1 <- factor(d$Month1,
                  levels = c("January","May", "June", "September", "October"),ordered = TRUE)

```


```{r Plot to have a look at the data (?)}
ggplot(d)+
  geom_point(aes(x=Date, y=Hg.DOC, shape=Lake, color = factor(Month1)), size=5)+  
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  scale_color_discrete("Seasons:", labels = c("Mar - May", "Jun - Aug",
                                              "Sep - Nov", "Dec - Feb"))

    ylim(0, -0.013)
    facet_grid(.~"%Y")


```

```{r #01 Barrplot, simple, FRACTIONS biodeg w errorbar. Is overall LMW more biodegradable than HMW?}

hg <- select(d, Fraction, Lake, MeHg.DOC)
head(hg)
hg <- na.omit(hg)
dx <- summarySE(hg, measurevar="MeHg.DOC", groupvar=c("Fraction", "Lake"))
  head(d)
head(dx)
dx
ggplot(dx) +
  geom_bar( aes(x=Fraction, y=MeHg.DOC, fill=factor(Fraction)), stat="identity", alpha=0.7)+ 
  geom_errorbar(aes(x=Fraction, ymin=MeHg.DOC-sd, ymax=MeHg.DOC+sd), width=0.2, colour="black", alpha=0.9, size=1)+
  facet_grid(.~Lake)
  ylim(0, -0.009)


```
```{r #011 STatistical testing}
 ny <- d %>% select(Lake, Fraction, Rel.rate.DOC)
  ny
  
  dx
  res.aov <- aov(MeHg.DOC ~ Lake, data = d)
  # Summary of the analysis
  summary(res.aov)
  TukeyHSD(res.aov)
  nyt
  t.test(y1,y2)
  
  t.test(d$Hg.DOC ~ d$Lake)
  
  head(nytt)
```


```{r #1) Barplot: seasonal, relative}
hg <- select(d, sUVa, Lake)
hg <- select(d, Month, Month1, Year, Rate.DOC.ave)
head(Hg)
hgC <- select(d, Hg, Hg.DOC, Lake, Fraction)
hg = na.omit(hg)
hgC = na.omit(hgC)
dx <- summarySE(, measurevar="Rate.DOC.ave", groupvar=c("Month"))
dx$miny <- dx$Rate.DOC.ave - dx$sd
dx$maxy <- dx$Rate.DOC.ave + dx$sd
head(dx)
summary(hg)

ggplot(d) + 
  geom_bar(aes(x=Fraction, y=Rate.DOC.ave, fill=factor(Fraction)), stat="identity")
    geom_errorbar(aes(x=Fraction, ymin=Rate.DOC.ave-sd, ymax=Rate.DOC.ave+sd),
                width=.2,                    
                position="identity", col="black")
  theme(axis.text.y = element_text(size= 20, colour="black"),
        axis.title.x = element_text(size = 20, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 24, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 26),
        axis.text.x = element_text(angle=0, vjust=1, size = 24, colour="black"),
        plot.title=element_text(size=30))+
  labs(title = "Hg/DOC", x = "", y = "Hg:DOC (ng/mg)")

#scale_fill_manual(values=c("#F8766D", "#7CAE00", "#00BFC4"))+
```


```{r #2) Old barplot: Rate.DOC per season, per fraction, per lake}
ggplot(dx, aes(x=factor(Fraction), y=SAR, group=factor(Fraction), position="dodge")) + 
  geom_bar(aes(x=factor(Fraction)), stat="summary", fun.y="mean")+
  geom_errorbar(aes(ymin=SAR-sd, ymax=SAR+sd),
                width=.2, colour="orange", size=1)+
  labs(y="SAR (Abs 254nm/Abs 400nm)")+
  ggtitle("Molecular size and aromaticicty")
  theme(axis.text.y = element_text(size= 14),
        axis.title.x = element_text(size = 18, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 18, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 18),
        axis.text.x = element_text(angle=0, hjust=1, vjust=0.5, size = 16),
        plot.title=element_text(size=20, face="bold"))+
  facet_grid(. ~ factor(Lake))
```
```{r # 3) BOX PLOT: TOtal}
head(d)
ggplot(d, aes(x=factor(Fraction), fill=factor(Fraction), y=Rel.rate.DOC, group=factor(Fraction))) + 
  geom_boxplot(size=1.2, width=1)+
  labs(y="Relative Rate of DOC decay/DOC (%)",
       x="")+
  theme(axis.text.x = element_text(size = 28, angle=30, hjust=1, vjust=1, colour="black"),
        axis.text.y = element_text(size= 20, colour="black"),
        axis.title.x = element_text(size = 24, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 26, margin=margin(0,20,0,0), colour="black"),
        strip.text.x = element_text(size = 30, face="plain"))+
  facet_grid(. ~ factor(Lake))+
    scale_fill_manual(values=c("#F8766D", "#7CAE00", "#00BFC4"))
  ylim(0, -0.013)

ggtitle("")
head(d)
axis.text.x = element_text(angle=90, hjust=1, vjust=0.5, size = 16)


```
#4) Correlation plots w biodegradation
#Correlation plots, LMW/HMW, and INlet/Outlet
```{r Correlation plots}
d <- subset(d, Lake == "Inlet"| Lake =="Outlet")
d <- subset(d, Fraction == "LMW"| Fraction =="HMW")
head(d)
#IMPORTANT!!! Correlation plot for biodeg and parameters
ggplot(d, aes(E2_E3, Hg.DOC))+
  geom_point(aes(colour=interaction(factor(Fraction)), shape=factor(Lake)), size=6)+
  geom_smooth(method = "lm", se = FALSE, colour="black")
  facet_grid(. ~ factor(Lake), scales="free")+
  xlim(4, 6)
  

ylim(-0.013, 0)
 theme(axis.text.y = element_text(size= 22),
        axis.title.x = element_text(size = 24, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 24, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 32),
        axis.text.x = element_text(angle=0, hjust=0.5, vjust=0, size = 24),
        plot.title=element_text(size=30, face="bold"),
        legend.text=element_text(size=26),
        legend.title = element_blank())+
    ylim(0, -0.013)
  labs(title = "DOM Size and Aromaticity", x = "sUVa (Abs254:DOC*100)", y = "Rate of DOC decay:DOC (/h)")
  ylim(0, -0.007)+
  scale_colour_manual(values=c("#F8766D", "#7CAE00", "#00BFC4"))+
  xlim(0, 8)

xlim(0,8)
scale_colour_manual(values=c("#fcbfba", "#a8e01d", "#10e2e8",
                             "#F8766D","#7CAE00", "#00BFC4"))+

  #IMPORTANT: PLOT for correlation between e.g. Hg with lake as facet and size colour coded
  ggplot(d, aes(sUVa, Hg))+
  geom_point(aes(colour=interaction(factor(Lake),factor(Fraction))), size=6)+
  geom_smooth(method = "lm", se = FALSE, colour="black")+
  facet_grid(. ~ factor(Lake))+
  theme(axis.text.y = element_text(size= 22),
        axis.title.x = element_text(size = 24, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 24, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 28),
        axis.text.x = element_text(angle=0, hjust=0.5, vjust=0, size = 22),
        plot.title=element_text(size=30, face="bold"),
        legend.text=element_blank(),
        legend.title = element_blank())+
  labs(title = "DOM Size and Aromaticity", x = "E2_E3 (Abs 254:365nm)", y = "THg:DOC (ng/mg)")+
  scale_colour_manual(values=c("#fcbfba", "#a8e01d", "#10e2e8",
                               "#F8766D","#7CAE00", "#00BFC4"))
facet_wrap(~variable, scales = "free_x")
head(d)
```

```{r Hg analysis}
# 5) Hg analysis
#Divide Hg on DOC
d <- transform(d, Hg.DOC = Hg / DOC)
d <- transform(d, MeHg.DOC = MeHg / DOC)
d
d <- subset(d, Fraction =="HMW")
d$Fraction <- factor(d$Fraction,
                     levels = c("LMW","HMW"),ordered = TRUE)
d$Lake <- factor(d$Lake,
                 levels = c("Inlet","Hypolimnion", "Outlet"),ordered = TRUE)
d$Month <- factor(d$Month,
                  levels = c("September_16", "January_17", "April_17", "June_17", "September_17", "May_18", "September_18", "October_18"),ordered = TRUE)
d<-d[(d$Hg< 8),]
d<-d[(d$SAR< 15),]

#dB <- subset(d, Fraction == "Bulk")
head(d)
d
```

```{r Barplot for Hg}

```

# 5.1( Barpoloy for Hg)
library(Rmisc)
library(ggplot2)

library(dplyr)
head(d)
d<-d[(d$Month=="September_16" | d$Month=="September_17" | d$Month=="April_17"),]

hg <- select(d, Rate.DOC.ave, Month, Lake, Fraction, DOC, Hg, Hg.DOC)
mehg <- select(d, Lake, MeHg)
head(mehg)
#plot(MeHg~DOC, data=mehg)
#fit <- lm(mehg$MeHg~ mehg$DOC)
#abline(fit, col="red")
#summary(fit)

mehg = na.omit(mehg)
hg = na.omit(hg)
#IMPORTANT: MEHG BARPLOT
dx <- summarySE(mehg, measurevar="MeHg", groupvar=c("Lake"))
head(dx)
d <- subset(d, Fraction == "LMW"| Fraction =="HMW")
head(dx)
ggplot(dx, aes(x=factor(Lake), y=MeHg, fill=Lake)) + 
  geom_bar(stat = "summary", fun.y = "mean", position="stack")+
  labs(y="MeHg (ng/L)",
       x="", title="MeHg")+
  theme(axis.text.x = element_text(size = 28, angle=30, hjust=1, vjust=1, colour="black"),
        axis.text.y = element_text(size= 22, colour="black"),
        axis.title.x = element_text(size = 20, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 24, margin=margin(0,20,0,0)),
        plot.title = element_text( size = 28))+
  #facet_grid(. ~ factor(Lake))+
  geom_errorbar(aes(ymin=MeHg-sd, ymax=MeHg+sd),width=.2, colour="black")+
  scale_fill_manual(values=c("#F8766D","#7CAE00", "#00BFC4"))


inlet <- subset(d, Lake == "Inlet")
hypo <- subset(d, Lake== "Hypolimnion")
outlet <- subset(d, Lake == "Outlet")

lmw <- subset(d, Fraction == "LMW")
hmw <- subset(d, Fraction == "HMW")
head(d)
#Is there a correlation?
plot(hypo$Hg.DOC~hypo$E2_E3)
fit <- lm(hypo$Hg.DOC~ hypo$E2_E3)
abline(fit, col="red")
summary(fit)

plot(HMW$Rate.DOC.ave~HMW$HI)

fit <- lm(outlet$Rate.DOC.ave~ outlet$HI)
#fit <- lm(LMW$Rate.DOC.ave~ LMW$sVISa+LMW$SAR+LMW$a)
abline(fit, col="red")
summary(fit)


#Correlation analysis
library(dplyr)
b <- select(ou, DOC, pH, Cond, Hg, MeHg, sUVa, sVISa, SAR, E2_E3, HI, Hg.DOC, MeHg.DOC)
library(Hmisc)
rcorr(as.matrix(b), type="pearson")
class(d)
library("PerformanceAnalytics")
chart.Correlation(b, histogram=TRUE, pch=19)
head(d)


#MAKKING THE CORRELATION ASSESSMENT
library(tidyverse)
library(ggpubr)
theme_set(theme_pubr())
ggplot(inl, aes(x = SAR, y = Hg.DOC)) +
  geom_point() +
  stat_smooth()+
  stat_smooth(method = lm)

model <- lm(Hg.DOC ~ sVISa, data = bu)
model  
summary(model)





#Divide into lake location
bu <- subset(d, Lake == "Inlet"| Lake =="Outlet")
outlet <- subset(d, Lake == "Outlet")
depth <- subset(d, Lake == "Hypolimnion")
#d <- subset(d, Lake == "Inlet"| Lake =="Outlet")
#Divide into DOM size fraction
LMW <- subset(d, Fraction == "LMW")
d <- subset(d, Fraction == "LMW"| Fraction =="HMW")
HMW <- subset(d, Fraction == "HMW")
Total <- subset(d, Fraction == "Bulk")
#SUbset by both fraction and location: to see seasonal variation
inl <- subset(d, Lake == "Inlet")
ou <- subset(d, Lake == "Outlet")
dep <- subset(HLMW, Lake == "Depth")
plot(inlet$Rate.DOC.ave~inlet$sUVa)

#Summarize data before plotting
library(Rmisc)
library(ggplot2)
dx <- summarySE(d, measurevar="Rate.DOC.ave", groupvar=c("Fraction"))
dx
DOC <- summarySE(d, measurevar="DOC", groupvar=c("Lake", "Fraction"))
DOC

nyt<-ny[!(ny$Fraction=="HMW"),]
nytt<-nyt[!(nyt$Lake=="Hypolimnion"),]
head(ny)
head(d)

#STATISTICAL TESTING (do not summarise first)
#Grubbs test
library(outliers)
grubbs.test(nytt$, type = 10, opposite = FALSE, two.sided = FALSE)


head(d)
d
library(tidyverse)
ny <- d %>% select(Lake, Fraction, Rel.rate.DOC)
ny
nyt<-d[(d$Fraction=="LMW"),]
res.aov <- aov(Rel.rate.DOC ~ Lake, data = nyt)
# Summary of the analysis
summary(res.aov)
TukeyHSD(res.aov)
nyt
t.test(y1,y2)

t.test(nytt$Rel.rate.DOC ~ nytt$Lake)
head(nytt)

#ORDERING the factor variables
d$Lake <- factor(d$Lake,
                 levels = c("Inlet","Hypolimnion", "Outlet"),ordered = TRUE)
d$Fraction <- factor(d$Fraction,
                     levels = c("LMW","HMW", "Total"),ordered = TRUE)




write.table(DOC, "//niva-of5/osl-userdata$/CBG/Documents/CLIMER/Data/Biodegradation/DOCsum", sep="\t")
par(mfrow=c(1,2)) 
head(LMW)

#Is there a correlation?
outlet<-outlet[(outlet$sUVa> 1),]
head(inlet)
plot(d$Rate.DOC.ave~d$sUVa)
plot(d$Rate.DOC.ave~d$sUVa)
fit <- lm(d$Rate.DOC.ave~ d$sUVa)
fit <- lm(outlet$Rate.DOC.ave~ outlet$HI)
#fit <- lm(LMW$Rate.DOC.ave~ LMW$sVISa+LMW$SAR+LMW$a)
abline(fit, col="red")
summary(fit)


#repeated from summary section above
dx <- summarySE(d, measurevar="Hg", groupvar=c("Month", "Lake", "Fraction"))
head(dx)

#ORDERING the factor variables
dx$Lake <- factor(dx$Lake,
                  levels = c("Inlet","Hypolimnion", "Outlet"),ordered = TRUE)
dx$Fraction <- factor(dx$Fraction,
                      levels = c("LMW","HMW", "Total"),ordered = TRUE)
dx$Month <- factor(dx$Month,
                   levels = c("September_16", "January_17", "April_17", "June_17", "September_17", 
                              "May_18", "September_18", "October_18"),ordered = TRUE)
head(dx)
#4) plot biodeg data seasonal w error bars and lines
pd <- position_dodge(0.1)
ggplot(dx, aes(x=factor(Month), y=Hg, color=factor(Fraction))) + 
  geom_line(aes(group=Fraction), position=pd) +
  geom_point(position=pd)+
  scale_color_manual(values=c("blue", "orange", "red"))+
  facet_grid(. ~ factor(Lake))+
  theme(axis.text.y = element_text(size= 14),
        axis.title.x = element_text(size = 18, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 18, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 18),
        axis.text.x = element_text(angle=90, hjust=1, vjust=0.5, size = 16),
        plot.title=element_text(size=20, face="bold"))
scale_x_discrete(breaks=1:26,
                 labels=c("September_16", "October_16", "November_16", "December_16",
                          "January_17", "February_17", "March_17", "April_17", "May_17",
                          "June_17", "July_17", "August_17", "September_17", "October_17",
                          "November_17", "December_17", "January_18", "February_18", "March_18", 
                          "April_18", "May_18","June_18", "July_18", "August_18", 
                          "September_18", "October_18" ))






#For ? se effekt av seson p? LMW og HMW sesongvariasjon
head(inl2)
ggplot(ou2, aes(x=factor(Month), y=Rate.DOC,  fill=Fraction)) + 
  geom_bar(stat="identity", position=position_dodge())+
  geom_errorbar(aes(ymin=Rate.DOC-se, ymax=Rate.DOC+se),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))


ggplot(depth, aes(x=factor(Fraction), y=Rate.DOC, group=factor(Fraction))) + 
  geom_boxplot()+
  facet_grid(. ~ factor(Month))+
  labs(y="Rate of O2 decay",
       x="DOM fraction")+
  ggtitle("Depth")+
  theme(axis.text.x = element_text(size = 16, face = "plain"),
        axis.text.y = element_text(size= 14),
        axis.title.x = element_text(size = 18, margin=margin(20,0,0,0)),
        axis.title.y = element_text(size = 18, margin=margin(0,20,0,0)),
        strip.text.x = element_text(size = 18),
        plot.title = element_text(size=30, face="bold", margin=margin(0,0,0,20)))

ggplot(lmw, aes(x=factor(Month), y=Rate.DOC, group=factor(Month), fill=Fraction) + 
         geom_bar(stat="identity")+
         geom_errorbar(aes(ymin=Rate.DOC-se, ymax=Rate.DOC+se),
                       width=.2,                    # Width of the error bars
                       position=position_dodge(.9))
       
       geom_boxplot()+
         facet_grid(. ~ factor(Lake))+
         labs(y="Rate of O2 decay",
              x="DOM fraction")+
         theme(axis.text.x = element_text(size = 16, face = "plain"),
               axis.text.y = element_text(size= 14),
               axis.title.x = element_text(size = 18, margin=margin(20,0,0,0)),
               axis.title.y = element_text(size = 18, margin=margin(0,20,0,0)),
               strip.text.x = element_text(size = 18))
  
