---
title: "220602_CLIMER_monitoringData"
author: "CBG"
date: "2 6 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The following file can not be found. 
```{r}
#setwd("//niva-of5/osl-userdata$/CBG/Documents/R/Projects/CLIMER TFF/CLIMER TFF")
#dx <- read.table("TOC_Hg_Abs_Inlet__Hyp_outlet_2016_2019_2.txt", sep = "\t", header = TRUE)
#found errors in the abs data at 365 nm. May also be errors at other wavelengths, for the inlet. Outlet seems ok and 254 nm ok
dx <- read.table("TOC_Hg_Abs_Inlet__Hyp_outlet_2016_2019_3.txt", sep = "\t", header = TRUE)

```

```{r}
library(ggplot2)
library(magrittr)
library(dplyr)
library(scales)
library(lubridate)
library(anytime) 
```


```{r}
#FIND variability in inlet vs outlet TOC
outlet <- subset(dx,  Station =="Outlet")
inlet <- subset(dx,  Station =="Inlet")
#dx<-dx[!(dx$Station=="Hypolimnion"),]

library(pastecs)
stat.desc(outlet)

dx$Date <- as.Date(dx$Date,format = "%d.%m.%Y")

dx$Year=format(as.Date(dx$Date, format="%d.%m.%Y"),"%Y")
dx$month_day=format(as.Date(dx$Date, format="%d.%m"),"%d.%m")
dx$month_day <- as.Date(dx$month_day,format = "%d.%m")


#dx$year <- tidyr::separate(dx, Date, c('year'), sep = "-",remove = FALSE)
#dx$year <- transform(dx,  year = format(Date, "%Y"))

#date = format(Date, "%d"), month = format(Date, "%m"),

head(dx)
str(dx)

#dx$labile_m = (dx$X285nm/dx$TOC_m)*100
#dx$labile_s = (dx$X285nm_s/dx$DOC_s)*100

dx$sUVa_m = (dx$X254nm/dx$TOC_m)*100
dx$sUVa_s = (dx$X254nm_s/dx$DOC_s)*100

#dx$sVISa_m = (dx$X400nm/dx$TOC_m)*1000
#dx$sVISa_s = (dx$X400nm_s/dx$DOC_s)*1000

#dx$SAR_m = (dx$X254nm/dx$X400nm)*1000
#dx$SAR_s = (dx$X400nm_s/dx$DOC_s)*1000

dx$THg_m = (dx$X254nm/dx$X365nm)
dx$THg_s = (dx$X254nm_s/dx$X365nm_s)

dx$THGTOC_m = dx$THg_m/dx$TOC_m
dx$THGDOC_s = dx$THg_s/dx$DOC_s

dx$MeHGTOC_m = dx$MeHg_m/dx$TOC_m
dx$MeHGDOC_s = dx$MeHg_s/dx$DOC_s

dx$MeHg_rel_m = (dx$MeHg_m/dx$THg_m)*100
summary(dx$MeHg_rel_m)
head(dx)
```

```{r Plotting monitoring data}
#PLOTTING with time on x-axis and year-facets
class(dx$Date)
class(dx$month_day)

#TOC
ggplot(dx)+
  geom_point(aes(x=month_day, y=TOC_m, shape=Station, colour=Year), size=5)+
  #geom_line(aes(x=month_day, y=TOC_m, colour=Station))+
  geom_smooth(aes(x=month_day, y=TOC_m, colour=Station), shape=16, size=1)+
  geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)
  
ggplot(dx)+
  geom_point(aes(x=month_day, y=TOC_m, colour=Station), shape=16, size=5)+
  #geom_line(aes(x=month_day, y=TOC_m, colour=Station))+
  geom_smooth(aes(x=month_day, y=TOC_m, colour=Station), shape=16, size=1)+
  geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)
  
  
#THg:TOC
ggplot(dx)+
  geom_point(aes(x=month_day, y=THGTOC_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=month_day, y=THGTOC_m, colour=Station), shape=16, size=2)+
  geom_point(aes(x=month_day, y=THGDOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "THg:TOC (ng/mg)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)  

#sUVa
ggplot(dx)+
  geom_point(aes(x=month_day, y=sUVa_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=month_day, y=sUVa_m, colour=Station), shape=16, size=2)+
  #geom_point(aes(x=month_day, y=sUVa_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
    labs(y = "sUVa-254nm (L/mg C m)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)  

#Hg
ggplot(dx)+
  geom_point(aes(x=month_day, y=THg_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=month_day, y=THg_m, colour=Station), shape=16, size=2)+
  geom_point(aes(x=month_day, y=THg_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
     labs(y = "THg (ng/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)  


#size
ggplot(dx)+
  geom_point(aes(x=Date, y=TOC_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=Date, y=TOC_m, colour=Station), shape=16, size=2)+
  #geom_point(aes(x=month_day, y=THg_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
     labs(y = "THg")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )
  facet_grid(.~ Year)  

ggplot(dx)+
  geom_point(aes(x=month_day, y=THg_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=month_day, y=THg_m, colour=Station), shape=16, size=2)+
  #geom_point(aes(x=month_day, y=THg_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
     labs(y = "THg")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)  

ggplot(dx)+
  geom_point(aes(x=month_day, y=SAR_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=month_day, y=SAR_m, colour=Station), shape=16, size=2)+
  #geom_point(aes(x=month_day, y=THg_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "SAR")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year) +
  ylim(7000, 9000)

, scales="free_x")
  ylim(4, 7)
      
  facet_grid(.~year, scales="free_x")
scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
 
  ggplot(dx)+
  geom_point(aes(x=Date, y=TOC_m, colour=Station), shape=16, size=5)+
  geom_smooth(aes(x=Date, y=TOC_m, colour=Station), shape=16, size=2)+
      scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  facet_grid(.~year(dx$Date), scales="free_x")+
  theme_bw()
  ylim(4, 7)

scale_y_continuous(limit=c(3.2, 4))
geom_point(aes(x=Date, y=labile_s, colour=Station), shape=17, size=5)+
head(dx)
```

something trange with THg, needs to investigae
```{r}
write.csv(dx, "climer checkin.csv")
```

21.06.2024 - test for significant difference between inlet and outlet in 
TOC, sUVa, THg, THg and THg-TOC
- first check if data is normally distrubuted (Shapiro-wilk test)
From the output, the p-value > 0.05 implying that the distribution of the data are not significantly different from normal distribution. In other words, we can assume the normality.
- then perform corrct test (t-test or Mann Whitney )

If the sample size is large enough (n > 30), we can ignore the distribution of the data and use parametric tests.
```{r normal ditribution and difference inlet/outlet}
library(dplyr)
library(ggpubr)

inlet_TOC <- subset(dx, Station == "Inlet")
outlet_TOC <- subset(dx, Station == "Outlet")
length(inlet_TOC$TOC)
length(outlet_TOC$TOC)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_TOC$TOC, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_TOC$TOC, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_TOC$TOC)
ggqqplot(outlet_TOC$TOC)

#Normality test
shapiro.test(inlet_TOC$TOC)
shapiro.test(outlet_TOC$TOC)

```

```{r sUVa normality test}
inlet_sUVa_m <- subset(dx, Station == "Inlet")
outlet_sUVa_m <- subset(dx, Station == "Outlet")
length(inlet_sUVa_m$sUVa_m)
length(outlet_sUVa_m$sUVa_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_sUVa_m$sUVa_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_sUVa_m$sUVa_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_sUVa_m$sUVa_m)
ggqqplot(outlet_sUVa_m$sUVa_m)

#Normality test
shapiro.test(inlet_sUVa_m$sUVa_m)
shapiro.test(outlet_sUVa_m$sUVa_m)
```

```{r E2_E3 normality test}

inlet_E2_E3_m <- subset(dx, Station == "Inlet")
outlet_E2_E3_m <- subset(dx, Station == "Outlet")
length(inlet_E2_E3_m$E2_E3_m)
length(outlet_E2_E3_m$E2_E3_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_E2_E3_m$E2_E3_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_E2_E3_m$E2_E3_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_E2_E3_m$E2_E3_m)
ggqqplot(outlet_E2_E3_m$E2_E3_m)

#Normality test
shapiro.test(inlet_E2_E3_m$E2_E3_m)
shapiro.test(outlet_E2_E3_m$E2_E3_m)
```

```{r THg normality test}

inlet_THg_m <- subset(dx, Station == "Inlet")
outlet_THg_m <- subset(dx, Station == "Outlet")
length(inlet_THg_m$THg_m)
length(outlet_THg_m$THg_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_THg_m$THg_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_THg_m$THg_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_THg_m$THg_m)
ggqqplot(outlet_THg_m$THg_m)

#Normality test
shapiro.test(inlet_THg_m$THg_m)
shapiro.test(outlet_THg_m$THg_m)
```

```{r THgTOC_TOC normality testing}

inlet_THGTOC_m <- subset(dx, Station == "Inlet")
outlet_THGTOC_m <- subset(dx, Station == "Outlet")
length(inlet_THGTOC_m$THGTOC_m)
length(outlet_THGTOC_m$THGTOC_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_THGTOC_m$THGTOC_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_THGTOC_m$THGTOC_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_THGTOC_m$THGTOC_m)
ggqqplot(outlet_THGTOC_m$THGTOC_m)

#Normality test
shapiro.test(inlet_THGTOC_m$THGTOC_m)
shapiro.test(outlet_THGTOC_m$THGTOC_m)
```



```{r}
library(rstatix)
# Pairwise comparisons

library("ggpubr")
class(HMW$sUVa)
ggboxplot(dx, x = "Station", y = "THGTOC_m", 
          color = "Station", palette = c("#00AFBB", "#E7B800"),
          ylab = "Weight", xlab = "Groups")


y16 <- subset(dx, Year== "2016")

pwc1 <- dx %>%
  pairwise_t_test(TOC_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc2 <- dx %>%
  pairwise_t_test(sUVa_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc3 <- dx %>%
  pairwise_t_test(E2_E3_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc4 <- dx %>%
  pairwise_t_test(THg_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc5 <- dx %>%
  pairwise_t_test(THGTOC_m ~ Station, paired=FALSE, alternative= "two.sided")

pwc1
pwc2
pwc3
pwc4
pwc5


pwc6 <- y16 %>%
  pairwise_t_test(sUVa_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc7 <- y16 %>%
  pairwise_t_test(E2_E3_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc6
pwc7
```


Correlation plots
```{r}
library("ggpubr")
head(dx)

#dy<-dx[!(dx$Year=="2016"),]

ggscatter(dx, x = "E2_E3_m", y = "THg_m", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="TOC (mg/L)" , ylab ="THg (ng/L)")
  facet_grid(.~Station)

ggscatter(dx, x ="sUVa_m",  y ="THg_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="sUVa" , ylab ="THg (ng/L)")+
  facet_grid(.~Station)


ggscatter(dx, x ="sUVa_m",  y ="THGTOC_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="" , ylab ="")+
  facet_grid(.~Station)

ggscatter(dx, x ="E2_E3_m",  y ="THGTOC_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="" , ylab ="")+
  facet_grid(.~Station)


    facet_grid(.~year(dx$Date), scales="free_x")
    
head(dx)
class(dx$TOC_m)
    #IMPORTANT!!! Correlation plot for Hg and DOM 
  ggplot(dx, aes(TOC_m, THg_m))+
      geom_point(size=6)
      geom_smooth(method = "lm", se = FALSE, colour="black")
      facet_grid(. ~ factor(Station), scales="free")
      xlim(4, 6)
      
      inlet <- subset(dx, Station == "Inlet")
      outlet <- subset(dx, Station == "Outlet")
    head(dx)
      #Correlation analysis
      library(dplyr)
      b <- select(dx, TOC_m, THg_m, MeHg_m, sUVa_m, sVISa_m, SAR_m, THg_m, labile_m, THGTOC_m, MeHGTOC_m)
      library(Hmisc)
      rcorr(as.matrix(b), type="pearson")
      class(d)
      library("PerformanceAnalytics")
      chart.Correlation(b, histogram=TRUE, pch=19)
      head(d)
      
      
      #MAKKING THE CORRELATION ASSESSMENT
      library(tidyverse)
      library(ggpubr)
      theme_set(theme_pubr())
      ggplot(outlet, aes(x = SAR_m, y = THGTOC_m)) +
        geom_point() +
        stat_smooth()+
        stat_smooth(method = lm)
      
model <- lm(THGTOC_m ~ sVISa_m, data = dx)
model  
summary(model)
```

## Sumamry table

```{r create summary table}
library(dplyr)
library(rbind)

dxy <- do.call(rbind.data.frame, dx)

#dx_sum <- dx[c("Station", "TOC_m", "THg_m", "sUVa_m", "THg_m")]
dx_sum <- select(dx, "Station", "TOC_m", "THg_m", "sUVa_m", "THg_m")

typeof(dx_sum)
class(dx_sum$TOC_m)




sum <- dx_sum %>% 
  group_by(Station) %>% 
  summarise(across(.cols  = where(is.numeric), .f = list(mean = mean, min = min, max = max, sd = sd), na.rm = TRUE))
  
sum_n <- dx_sum %>% 
  group_by(Station) %>% 
  length(across(.cols  = where(is.numeric)))

write.csv(sum, "summary monitoing.csv")

```

  
#geom_point(aes(x=Date, y=sUVa_s, colour=Station), size=5, shape=17)+
geom_smooth(aes(x=Date, y=sUVa_s, colour=Station), size=2)
#Correlation plots
library("ggpubr")
head(dx)
ggscatter(dx, x = "sUVa_m", y = "sVISa_m", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")
  facet_grid(.~Station)
    facet_grid(.~year(dx$Date), scales="free_x")
    
head(dx)
    #IMPORTANT!!! Correlation plot for Hg and DOM 
    ggplot(inlet, aes(labile_m, THGTOC_m))+
      geom_point(size=6)+
      geom_smooth(method = "lm", se = FALSE, colour="black")
      facet_grid(. ~ factor(Station), scales="free")
      xlim(4, 6)
      
      inlet <- subset(dx, Station == "Inlet")
      outlet <- subset(dx, Station == "Outlet")
    head(dx)
      #Correlation analysis
      library(dplyr)
      b <- select(dx, TOC_m, THg_m, MeHg_m, sUVa_m, sVISa_m, SAR_m, THg_m, labile_m, THGTOC_m, MeHGTOC_m)
      library(Hmisc)
      rcorr(as.matrix(b), type="pearson")
      class(d)
      library("PerformanceAnalytics")
      chart.Correlation(b, histogram=TRUE, pch=19)
      head(d)
      
      
      #MAKKING THE CORRELATION ASSESSMENT
      library(tidyverse)
      library(ggpubr)
      theme_set(theme_pubr())
      ggplot(outlet, aes(x = SAR_m, y = THGTOC_m)) +
        geom_point() +
        stat_smooth()+
        stat_smooth(method = lm)
      
model <- lm(THGTOC_m ~ sVISa_m, data = dx)
model  
summary(model)



#STATISTICAL TESTING
res.aov <- aov(THGTOC_m ~ Station, data = dx)
# Summary of the analysis
summary(res.aov)
TukeyHSD(res.aov)







#Correlation with SENSOR LAKE DATA
dy <- read.table("Langtjern_weather.txt", sep = "\t", header = TRUE) #load data
dy$Date <- as.Date(dy$SampleDate,format = "%d.%m.%Y") #convert to date column
df <- dy[ -c(1:5) ] #removed uneccesary columns
head(df)

#to make monthly summaries
dff = df %>%
  mutate(month = format(Date, "%m"), year = format(Date, "%Y")) %>%
  group_by(month, year) %>%
  summarise(sun = mean(Straaling),
            airT = mean(AirTemp),
            Prec = mean(Precip),
            Hum = mean(Humidity))

plot(dff)
head(dff)
#Aggreger per dag
library(dplyr)
(out <- df %>%
    group_by(Date) %>%
    summarize(Sun=sum(Straaling), 
              Precip=sum(Precip),
              Precip_ave=mean(Precip),
              AirTemp=mean(AirTemp),
              Humidity = mean(Humidity)))

head(df)

#PLOTTING with time on x-axis and year-facets
ggplot(out)+
  geom_point(aes(x=Date, y=Precip), shape=16, size=5)
  geom_smooth(aes(x=Date, y=Precip), shape=16, size=2)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  facet_grid(.~year(dx$Date), scales="free_x")+
  theme_bw()
ylim(2.5, 6)
scale_y_continuous(limit=c(3, 4))
geom_point(aes(x=Date, y=labile_s, colour=Station), shape=17, size=5)+
  head(dx)






ou <- head(out)
dfff <- na.omit(dff)
head(dfff)
ou = dfff
nrow(sun)
head(ou)
#Make indivudual datafiles as ts with correct frequency
sun <- ts(ou$sun, frequency=12, start=2016, end=2019)
pre <- ts(ou$Prec, frequency=12, start=2016, end=2019)
airT <- ts(ou$airT, frequency=12, start=2016, end=2019)
Humi <- ts(ou$Hum, frequency=12, start=2016, end=2019)

plot(sunny)
sunny <- decompose(sun)
precip <- decompose(pre)
AirTemp <- decompose(airT)
Humidity <- decompose(Humi)
plot(Humidity)
plot(AirTemp)
plot(precip)
plot(sunny)
ts.plot(pre)

#then YOU CAN SUBTRACT E.G. THE SEAOSNAL COMPONENT FROM THE DATA
birthsSeasonAdj <- sun - sunny$seasonal- sunny$random
plot.ts(birthsSeasonAdj)

#Make correspodning individual ts with correct frequency
#here we have inlet and outlet. can look at these individually and at the difference bwteen the two
newdata <- subset(dx,  Station =="Outlet")
sUVa <- na.omit(newdata$sUVa_m)
sUVaTS <- ts(sUVa, frequency=365, start=2016, end=2019)

decomsUVa <- decompose(sUVaTS)
plot(decomsUVa)

suva <- sUVaTS - decomsUVa$seasonal - decomsUVa$random
plot.ts(suva)
plot(suva~birthsSeasonAdj)

abline(lm(suva~birthsSeasonAdj))

ab <- lm(suva~birthsSeasonAdj)
summary(ab)

#What about doing monthly aggregates???








#PLoT værdata med eventuelt langtjern data
ggplot()+
  geom_smooth(dx, mapping=aes(x=Date, y=Precip), shape=16, size=2)+
  geom_point(dx, mapping=aes(x=Date, y=TOC_m, colour=Station), shape=16, size=3)
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  scale_y_continuous(name = "Radiation", 
                     sec.axis = sec_axis(~.*10, name = "TOC"))
  facet_grid(.~year, scales="free_x")+
  theme_bw()
  

linearMod <- lm(_m ~ TOC_m, data=dx)
print(linearMod)
summary(linearMod)
