---
title: "220602_CLIMER_monitoringData"
author: "CBG"
date: "2 6 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following file can not be found. 
```{r load data}
dx_o <- read.table("TOC_Hg_Abs_Inlet__Hyp_outlet_2016_2019_4.txt", sep = "\t", header = TRUE)

library(plyr)
dx_o$Station <- revalue(dx_o$Station, c("Langtjern, utl??p" = "Outlet","Innl??psbekk ??st" = "Inlet"))
```

```{r}
#install.packages("tidyverse")
library(ggplot2)
library(magrittr)
library(plyr)
library(dplyr)
library(scales)
library(lubridate)
library(anytime) 
library(tidyr)
library(pastecs)
library(ggpubr)
library(rstatix)
library(Hmisc)
library("PerformanceAnalytics")
library(tidyverse)
```

Extra
```{r}
#mA1 <- gamm(
#  TOC_m ~ Station + s(doy, bs = "cc", k = 24),
#  data = dx_G,
#  method = "REML",
#  #control = ctrl, #supposed to help, common for environmental data. Does not seem to make a difference
#  correlation = nlme::corCAR1(form = ~ t | Station)  # <-- grouping added
#)

#The coefficient for SiteOutlet tests whether the Outlet mean differs from Inlet after accounting for seasonality and autocorrelation.
# QB: Is one site consistently greater across the year?”
#mA2 <- gamm(
#  TOC_m ~ Station + s(doy, bs = "cc", k = 24) + s(t, k = 10, bs = "cs"),
#  data = dx_G,
#  method = "REML",
#  correlation = nlme::corCAR1(form = ~ t | Station),  # <-- grouping added
#  #control = ctrl
#)
#summary(mA1$gam)

#Site coefficient now answers: consistently higher/lower, averaged over the period, after season and long‑term trend are removed.
#QC: Do the temporal patterns differ between sites?

#mB1 <- gamm(
#  TOC_m ~ Site + s(doy, bs = "cc", k = 24, by = Site)+ s(t, k = 10, bs = "cs"),
#  data = dx_G,
#  method = "REML",
#  correlation = nlme::corCAR1(form = ~ t | Site),  # <-- grouping added
#  #control = ctrl
#)
#summary(mB1$gam)
```

Adding extended abs-data
```{r}
#load
abs <- read.table("263001_Extended abs data.txt", sep = "\t", header = TRUE)

#from long to wide
abs_wide <- abs %>%
  pivot_wider(
    names_from = wavelength,  # The column whose values become new column names
    values_from = abs_per_cm # The column whose values fill the new columns
  )

abs_wide$Date <- as.Date(abs_wide$sample_date,format = "%d.%m.%Y")
dx_o$Date <- as.Date(dx_o$Date,format = "%d.%m.%Y")

dxx <- left_join(dx_o, abs_wide, by = c("Station.ID", "Date"))

dxx$Date <- as.Date(dxx$Date,format = "%d.%m.%Y")
dxx$Year=format(as.Date(dxx$Date, format="%d.%m.%Y"),"%Y")
dxx$month_day=format(as.Date(dxx$Date, format="%d.%m"),"%d.%m")
dxx$month_day <- format(as.Date(dxx$Date, format="%d.%m"),"%d.%m")

dxx$month =format(as.Date(dxx$Date, format="%d.%m.%Y"),"%m")

dx <- subset(dxx, Year %in% c(2016, 2017, 2018, 2019, 2020, 2021))
class(dx$Date)
```

```{r}
#FIND variability in inlet vs outlet TOC

#library(pastecs)
#stat.desc(outlet)

dx$sUVa_m = (dx$"254"/dx$TOC_m)*100 #with new abs data from Areti
dx$sUVa_s = (dx$X254nm_s/dx$DOC_s)*100

dx$THGTOC_m2 = dx$THg_m/dx$TOC_m
dx$THGDOC_s2 = dx$THg_s/dx$DOC_s
```

Using GAMM to look for significant difference between inlet and outlet
```{r inlet versus outlet}
library(mgcv)   # gam/gamm/bam
library(nlme)   # correlation structures used by gamm()
library(gratia) # nicer plots/diagnostics for mgcv

# 1) TOC
dxTOC <- dx[!is.na(dx$TOC_m), ]
# Add seasonality and numeric time - THIS IS THE ONE!!!!!!!!!!!
dxTOC <- dxTOC[order(dxTOC$Date), ]
dxTOC$doy <- yday(dxTOC$Date)          # day of year for seasonal smooth
dxTOC$time_num <- as.numeric(dxTOC$Date) / (60*60*24)  # days since epoch
dxTOC$lnTOC <- log(dxTOC$TOC_m)

m1 <- gamm(
  lnTOC ~ Station +
    s(time_num, k = 40) +            # long-term nonlinear trend
    s(doy, bs = "cc", k = 30),       # seasonal cycle, cyclic smooth
  random = list(Date = ~1),
  correlation = corAR1(form = ~ time_num| Station),
  data = dxTOC
)

summary(m1$gam)
gam.check(m1$gam)
summary(m1$lme)
acf(resid(m1$lme, type = "normalized"), lag.max = 24)

anova(m1$lme)

# 2) sUVa
dxsUVa <- dx[!is.na(dx$sUVa_m), ]

# Add seasonality and numeric time - THIS IS THE ONE!!!!!!!!!!!
dxsUVa <- dxsUVa[order(dxsUVa$Date), ]
dxsUVa$doy <- yday(dxsUVa$Date)          # day of year for seasonal smooth
dxsUVa$time_num <- as.numeric(dxsUVa$Date) / (60*60*24)  # days since epoch
dxsUVa$lnsUVa <- log(dxsUVa$sUVa_m)

m2 <- gamm(
  lnsUVa ~ Station +
    s(time_num, k = 40) +            # long-term nonlinear trend
    s(doy, bs = "cc", k = 30),       # seasonal cycle, cyclic smooth
  random = list(Date = ~1),
  correlation = corAR1(form = ~ time_num| Station),
  data = dxsUVa
)

summary(m2$gam)
gam.check(m2$gam)
summary(m2$lme)
anova(m2$lme)

# 3) Hg
dxHg <- dx[!is.na(dx$THg_m), ]

# Add seasonality and numeric time - THIS IS THE ONE!!!!!!!!!!!
dxHg <- dxHg[order(dxHg$Date), ]
dxHg$doy <- yday(dxHg$Date)          # day of year for seasonal smooth
dxHg$time_num <- as.numeric(dxHg$Date) / (60*60*24)  # days since epoch
dxHg$lnHg <- log(dxHg$THg_m)

m3 <- gamm(
  lnHg ~ Station +
    s(time_num, k = 40) +            # long-term nonlinear trend
    s(doy, bs = "cc", k = 30),       # seasonal cycle, cyclic smooth
  random = list(Date = ~1),
  correlation = corAR1(form = ~ time_num| Station),
  data = dxHg
)

summary(m3$gam)
gam.check(m3$gam)
summary(m3$lme)

anova(m3$lme)

# 4) Hg:DOC
dxHgDOC <- dx[!is.na(dx$THGTOC_m2), ]

# Add seasonality and numeric time - THIS IS THE ONE!!!!!!!!!!!
dxHgDOC <- dxHgDOC[order(dxHgDOC$Date), ]
dxHgDOC$doy <- yday(dxHgDOC$Date)          # day of year for seasonal smooth
dxHgDOC$time_num <- as.numeric(dxHgDOC$Date) / (60*60*24)  # days since epoch
dxHgDOC$lnHgDOC <- log(dxHgDOC$THg_m)

m4 <- gamm(
  lnHgDOC ~ Station +
    s(time_num, k = 40) +            # long-term nonlinear trend
    s(doy, bs = "cc", k = 30),       # seasonal cycle, cyclic smooth
  random = list(Date = ~1),
  correlation = corAR1(form = ~ time_num| Station),
  data = dxHgDOC
)

summary(m4$gam)
gam.check(m4$gam)
summary(m4$lme)
anova(m4$lme)

# 6) Concurvity
concurvity(m3$gam, full = TRUE)














#DIAGNOSTICS
# Smooths (publication-quality)
draw(mB2$gam)             # or draw(mB1$gam), draw(mB2$gam)

# Residual autocorrelation (should be ~white noise)
acf(resid(mB2$lme, type = "normalized"), lag.max = 24)

# Residual checks (normality, heteroscedasticity, fitted vs residuals)
plot(mB2$lme)

# Basis dimension tests (interpret cautiously for gamm)
gam.check(mB2$gam)

# Optional: Compare models
#AIC(mA1$lme, mA2$lme, mB1$lme, mB2$lme)
```

```{r}
#extra old

#1) TOC
#remove NAs from student entries
dxc1 <- dx_G[!is.na(dx_G$TOC_m), ]

# QA: are the sites different on average?
ctrl <- nlme::lmeControl(niterEM = 60, msMaxIter = 200, maxIter = 100,
                         opt = "optim", msVerbose = FALSE)

#site specific short and long term time trends. USE THIS!
mB2 <- gamm(
  lnTOC ~ Site +
        s(doy, bs = "cc", k = 24, by = Site) +
        s(t,   k = 10, bs = "cs", by = Site),
  data = GdxYY,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t| Site),
  control = ctrl
)
summary(mB2$gam)
gam.check(mB2$gam)
acf(resid(mB2$lme, type = "normalized"), lag.max = 24)


# 2) sUVa
#remove NAs from student entries
dxc1 <- dx[!is.na(dx$sUVa_m), ]

dx_G1 <- dxc1 %>%
  mutate(
    Site = droplevels(factor(Station, levels = c("Inlet","Outlet"))),
    Date = as.Date(Date),
    doy  = yday(Date),                     # 1..365
    t    = as.numeric(Date - min(Date))    # days since start (continuous)
  ) %>%
  arrange(Station, Date)

#site specific short and long term time trends. USE THIS!
mB2 <- gamm(
  TOC_m ~ Site +
        s(doy, bs = "cc", k = 24, by = Site) +
        s(t,   k = 10, bs = "cs", by = Site),
  data = dx_G,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t| Site),
  #control = ctrl
)
summary(mB2$gam)
gam.check(mB2$gam)


#If the s(doy, by=Site) smooths are significantly different (or clearly different when plotted), inlet and outlet do not share the same seasonal cycle.

dxTOC_ <-dxTOC  %>%
  arrange(Station, Date) %>%
  mutate(
    doy = yday(Date),
    y = year(Date),
    t = as.numeric(difftime(Date, min(Date), units = "days")),
    lnHg = log(THg_m),               # often stabilizes variance; check first
    lnTOC = log(TOC_m),          # log1p if zeros are possible
    sUVa_sc = scale(sUVa_m),       # standardize for interpretability
    site = factor(Station)
  )


m_simple <- gamm(
  lnTOC ~ Station + s(doy, bs="cc", k=24), #+ s(t, bs="cs", k=10),
  data = dxTOC_,
  correlation = corCAR1(form = ~ t|Station),
  method = "REML"
)
summary(m_simple$gam)

gam.check(m_simple$gam)

acf(resid(m_simple$lme, type = "normalized"), lag.max = 24)

mB2 <- gamm(
  lnTOC ~ Station +
        s(doy, bs = "cc", k = 24, by = Station) +
        s(t,   k = 10, bs = "cs", by = Station),
  data = dxTOC_,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t| Station)
  #control = ctrl
)


dx_G1 <- dxc1 %>%
  mutate(
    Site = droplevels(factor(Station, levels = c("Inlet","Outlet"))),
    Date = as.Date(Date),
    doy  = yday(Date),                     # 1..365
    t    = as.numeric(Date - min(Date))    # days since start (continuous)
  ) %>%
  arrange(Station, Date)
```

##GAMM diagbnostics
```{r}
# 1) Smooths & significance
summary(m1_gamm_inl$gam)
plot(m1_gamm_inl$gam, pages = 1, shade = TRUE)

# 2)  Residuals vs Fitted
plot(m1_gamm_inl$lme)

# 3) Normality of Residuals
gam.check(m1_gamm_inl$gam)

# 4) Residual versus time
plot(resid(m1_gamm_inl$lme), type="l")

# 5 Autocorrelation Check (ACF of residuals). Residual correlation
acf(resid(m1_gamm_inl$lme, type = "normalized"), lag.max = 24)

# 6) Concurvity
concurvity(m1_gamm_inl$gam, full = TRUE)

# Summary of mixed model (AR1, random effects)
summary(m1_gamm_inl$lme)

```


Alt1 plot seasonal data (old)
```{r}
#TOC
ggplot(subset(dx, !is.na(TOC_m)))+
  geom_point(aes(x=month_day, y=TOC_m, shape=Station, colour=Year), size=5)+
  #geom_line(aes(x=month_day, y=TOC_m, colour=Station))+
  geom_smooth(aes(x=month_day, y=TOC_m, colour=Station), shape=16, size=1)+
  geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)

#Hg
ggplot(subset(dx, !is.na(THg_m)))+
  geom_point(aes(x=month_day, y=THg_m, shape=Station, colour=Station), size=4)+
  #geom_line(aes(x=month_day, y=TOC_m, colour=Station))+
  geom_smooth(aes(x=month_day, y=THg_m, colour=Station), shape=16, size=1)+
  geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year)

ggsave(filename = "output/TOC_m.png",
       plot = TOC,
       width = 11, height = 7, units = "in")

#THg:TOC
ggplot(data=subset(dx, !is.na(THg_m)))+
  geom_point(aes(x=month_day, y=as.numeric(sUVa_m), colour=Station, shape=Station), size=4.5)+
  geom_smooth(aes(x=month_day, y=as.numeric(sUVa_m), colour=Station), shape=16, size=1.2)+
  #geom_point(aes(x=month_day, y=as.numeric(THg_m), colour=Station), shape=17, size=4.5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "THg (ng/L)")+
   theme(
    legend.text = element_text(size = 16),
    legend.title = element_blank(),
    axis.text.x = element_text(size=16, angle = 45, hjust = 1, colour="black"),
    axis.text.y = element_text(size=17, colour="black"),
    strip.text = element_text(size = 17),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), size=17, colour="black")
  )+
  facet_grid(.~ Year)+
  scale_colour_manual(values = c("#0072B2", "#CC79A7"))


ggsave(filename = "output/Hg_DOC_m.png",
       plot = Hg_DOC,
       width = 11, height = 7, units = "in")

#sUVa
ggplot(dx)+
  geom_point(aes(x=month_day, y=sUVa_m, colour=Station), shape=16, size=4.5)+
  geom_smooth(aes(x=month_day, y=sUVa_m, colour=Station), shape=16, size=1.2)+
  #geom_point(aes(x=month_day, y=sUVa_s, colour=Station), shape=17, size=4.5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
    theme_bw()+
  labs(y = "sUVa-254nm (L/mg C m)")+
  theme(
    legend.text = element_text(size = 16),
    legend.title = element_blank(),
    axis.text.x = element_text(size=16, angle = 45, hjust = 1, colour="black"),
    axis.text.y = element_text(size=17, colour="black"),
    strip.text = element_text(size = 17),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), size=17, colour="black")
  )+
  facet_grid(.~ Year)+
  scale_colour_manual(values = c("#0072B2", "#CC79A7"))

ggsave(filename = "output/sUVa_m.png",
       plot = sUVa,
       width = 11, height = 7, units = "in")
```

```{r Monitoing inlet and outlet months}
class(dx$month_day)
dx$month_day <- as.Date(dx$month_day,format = "%d.%m")

#ggplot(subset(dx, Year %in% c(2016, 2017, 2018, 2019, 2020, 2021)))+
ggplot(dx)+
  geom_point(aes(x=month_day, y=as.numeric(sUVa_m), colour=Station, shape=Station), size=4.5)+
  geom_smooth(aes(x=month_day, y=as.numeric(sUVa_m), colour=Station), size=1.2)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("1 month"))+
  theme_bw()+
  #labs(y = "DOC (mg/L)")+
  labs(y = "sUVa254 (L/ mg C m) ")+
  #labs(y = "Hg (ng/L) ")+
  #labs(y = "Hg:DOC (ng/mg) ")+
   theme(
    legend.text = element_blank(),#size 16
    legend.title = element_blank(),
    axis.text.x = element_text(size=16, angle = 45, hjust = 1, colour="black"),
    axis.text.y = element_text(size=17, colour="black"),
    strip.text = element_text(size = 17),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), size=17, colour="black")
  )+
    scale_colour_manual(values = c("#0072B2", "#CC79A7"))+
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  ylim(3.5, 5.3) #only use this for suva to get space for the figure labelling

  #facet_grid(.~ Year)+
  #scale_colour_manual(values = c("#0072B2", "#CC79A7"))


ggsave(filename = "output/Hg_DOC_m.png",
       plot = Hg_DOC,
       width = 11, height = 7, units = "in")


#LOOk at difference between inlet and outlet in suva
dx$Dates = dx$Date
inl <- subset(dx, Station =="Inlet")
out <- subset(dx, Station =="Outlet")

both <- merge(inl, out, by="Dates")
both$diff <- (both$sUVa_m.x - both$sUVa_m.y)
both$diff2 <- (both$E2E3_m.x - both$E2E3_m.y)
both$diff3 <- (both$THGTOC_m2.x - both$THGTOC_m2.y)
both$diff4 <- (both$THg_m.x - both$THg_m.y)
both$THg_m.y
typeof(both$sUVa_m.x)

bothy <- subset(both, !Year.x == 2016)

ggplot(bothy)+
  #geom_point(aes(x=month_day.x, y=diff, colour=Year.x), shape=16, size=2)+
  geom_point(aes(x=month_day.x, y=diff3, colour=Year.x), size=2)+
  geom_smooth(aes(x=month_day.x, y=diff3), size=1.2)+
  #geom_point(aes(x=month_day, y=sUVa_s, colour=Station), shape=17, size=4.5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
    theme_bw()+
  labs(y = "Inlet - Outlet (THg)")+
  theme(
    legend.text = element_text(size = 16),
    legend.title = element_blank(),
    axis.text.x = element_text(size=16, angle = 45, hjust = 1, colour="black"),
    axis.text.y = element_text(size=17, colour="black"),
    strip.text = element_text(size = 17),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), size=17, colour="black")
  )+
  geom_hline(yintercept = 0)
  
ggsave(filename = "output/sUVa_m.png",
       plot = sUVa,
       width = 11, height = 7, units = "in")
```

something trange with THg, needs to investigae
```{r}
write.csv(dx, "climer checkin.csv")
```

21.06.2024 - test for significant difference between inlet and outlet in 
TOC, sUVa, THg, THg and THg-TOC
- first check if data is normally distrubuted (Shapiro-wilk test)
From the output, the p-value > 0.05 implying that the distribution of the data are not significantly different from normal distribution. In other words, we can assume the normality.
- then perform corrct test (t-test or Mann Whitney )

If the sample size is large enough (n > 30), we can ignore the distribution of the data and use parametric tests.
```{r normal ditribution and difference inlet/outlet}
#library(dplyr)
#library(ggpubr)

inlet_TOC <- subset(dx, Station == "Inlet")
outlet_TOC <- subset(dx, Station == "Outlet")
length(inlet_TOC$TOC)
length(outlet_TOC$TOC)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_TOC$TOC, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_TOC$TOC, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_TOC$TOC)
ggqqplot(outlet_TOC$TOC)

#Normality test
shapiro.test(inlet_TOC$TOC)
shapiro.test(outlet_TOC$TOC)

```

```{r sUVa normality test}
inlet_sUVa_m <- subset(dx, Station == "Inlet")
outlet_sUVa_m <- subset(dx, Station == "Outlet")
length(inlet_sUVa_m$sUVa_m)
length(outlet_sUVa_m$sUVa_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_sUVa_m$sUVa_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_sUVa_m$sUVa_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_sUVa_m$sUVa_m)
ggqqplot(outlet_sUVa_m$sUVa_m)

#Normality test
shapiro.test(inlet_sUVa_m$sUVa_m)
shapiro.test(outlet_sUVa_m$sUVa_m)
```

```{r THg normality test}
inlet_THg_m <- subset(dx, Station == "Inlet")
outlet_THg_m <- subset(dx, Station == "Outlet")
length(inlet_THg_m$THg_m)
length(outlet_THg_m$THg_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_THg_m$THg_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_THg_m$THg_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_THg_m$THg_m)
ggqqplot(outlet_THg_m$THg_m)

#Normality test
shapiro.test(inlet_THg_m$THg_m)
shapiro.test(outlet_THg_m$THg_m)
```

```{r THgTOC_TOC normality testing}
inlet_THGTOC_m <- subset(dx, Station == "Inlet")
outlet_THGTOC_m <- subset(dx, Station == "Outlet")
length(inlet_THGTOC_m$THGTOC_m)
length(outlet_THGTOC_m$THGTOC_m)

#Visual inspection: density plot and QQ plot
ggdensity(inlet_THGTOC_m$THGTOC_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggdensity(outlet_THGTOC_m$THGTOC_m, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")

ggqqplot(inlet_THGTOC_m$THGTOC_m)
ggqqplot(outlet_THGTOC_m$THGTOC_m)

#Normality test
shapiro.test(inlet_THGTOC_m$THGTOC_m)
shapiro.test(outlet_THGTOC_m$THGTOC_m)
```



Correlation plots
```{r}
library("ggpubr")
head(dx)

inelt <- subset(dx, Station =="Inlet")

dx <- dx %>%
  mutate(
    year = year(Date),   # Extracts the year
    month = month(Date), # Extracts the month as a number
    day = day(Date)      # Extracts the day of the month
  )

print(unique(dx$month))

dx2 <- dx %>%
  mutate(
    season = case_when(
      month == 12 ~ "Winter",
      month == 1 ~ "Winter",
      month == 2 ~ "Winter",
      month == 3 ~ "Spring",
      month == 4 ~ "Spring",
      month == 5 ~ "Spring",
      month == 6 ~ "Summer",
      month == 7 ~ "Summer",
      month == 8 ~ "Summer",
      month == 9 ~ "Autumn",
      month == 10 ~ "Autumn",
      month == 11 ~ "Autumn"
      # This line catches all remaining values
    )
  )

class(dx2$month)

#remove months of april and may to see if this improves correlations. this is for INLET. Month =4 and 5
# Values to remove
s_melt <- c(4, 5) #valid for inlet
#s_melt2 <- c(4, 5, 6, 7) #testing for outlet. Did not help!
s_melt2 <- c( 5,6)

# Remove rows where the city is in the cities_to_remove vector
filtered_dx2 <- dx2[!(dx2$month %in% s_melt), ]


#colour-code by month
ggplot(filtered_dx2)+
  geom_point(aes(x =(sUVa_m), y = (THg_m), colour=as.factor(month), shape=season), size=3)+
  facet_grid(.~Station)

ggplot(dx2)+
  geom_point(aes(x =(TOC_m), y = (THg_m), colour=as.factor(season)))+
  facet_grid(.~Station)


ggscatter(filtered_dx2, x = "TOC_m", y = "THg_m", #why no correlation to suva? greater seasonality in suva than TOC?
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="TOC (mg/L)" , ylab ="THg (ng/L)")+
          facet_grid(.~Station)

ggscatter(dx2, x ="TOC_m",  y ="THg_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="sUVa" , ylab ="THg (ng/L)")+
  facet_grid(.~Station)


ggscatter(dx, x ="sUVa_m",  y ="THGTOC_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="" , ylab ="")+
  facet_grid(.~Station)

ggscatter(dx, x ="E2_E3_m",  y ="THGTOC_m",  
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="" , ylab ="")+
  facet_grid(.~Station)


# facet_grid(.~year(dx$Date), scales="free_x")
    

#CONTINUING previous plots
    
head(dx)
class(dx$TOC_m)
    #IMPORTANT!!! Correlation plot for Hg and DOM 
  ggplot(dx, aes(TOC_m, THg_m))+
      geom_point(size=6)+
      geom_smooth(method = "lm", se = FALSE, colour="black")
      facet_grid(. ~ factor(Station), scales="free")
      xlim(4, 6)
      
      inlet <- subset(dx, Station == "Inlet")
      outlet <- subset(dx, Station == "Outlet")
    head(dx)
      #Correlation analysis
      #library(dplyr)
      b <- select(dx, TOC_m, THg_m, MeHg_m, sUVa_m, sVISa_m, SAR_m, THg_m, labile_m, THGTOC_m, MeHGTOC_m)
      #library(Hmisc)
      rcorr(as.matrix(b), type="pearson")
      class(d)
      #library("PerformanceAnalytics")
      chart.Correlation(b, histogram=TRUE, pch=19)
      head(d)
      
      
      #MAKKING THE CORRELATION ASSESSMENT
      #library(tidyverse)
      #library(ggpubr)
      theme_set(theme_pubr())
      ggplot(outlet, aes(x = SAR_m, y = THGTOC_m)) +
        geom_point() +
        stat_smooth()+
        stat_smooth(method = lm)
      
model <- lm(THGTOC_m ~ sVISa_m, data = dx)
model  
summary(model)
```

Upload and fix discharge data
- add to dx_ts?
```{r}
dx_ts <- dx[,c(3:6, 19, 40)]

dis <- read.table("Langtjern_discharge.txt", sep = "\t", header = TRUE)
dis$Date <- as.Date(dis$Date,format = "%d/%m/%Y")

#from wide to long
dis_long <- gather(dis, Station, Disch, LAE01.mm:LAE03.mm, factor_key=TRUE)
#rename station to inlet and outlet
dis_long$Station <- revalue(dis_long$Station, c("LAE01.mm" = "Outlet", "LAE03.mm" = "Inlet"))
head(dis_long)
head(dx_ts)

yp <- left_join(dx_ts, dis_long, by = c("Station", "Date"))
yp$Year=format(as.Date(yp$Date, format="%d.%m.%Y"),"%Y")

```

## t-test 30.03.2026

```{r}
library(rstatix)
# Pairwise comparisons

#suva not significantly different inlet and outlet, but what about during summer only?
# Values to keep
summer <- c( 6, 7, 8)

# Remove rows where the city is in the cities_to_remove vector
filtered_dx2 <- dx2[(dx2$month %in% summer), ]


library("ggpubr")
class(HMW$sUVa)
ggboxplot(yp, x = "Station", y = "TOC_m", 
          color = "Station", palette = c("#00AFBB", "#E7B800"),
          ylab = "Weight", xlab = "Groups")


yp1 <- yp[!is.na(yp$TOC_m),]
pwc1 <- yp1 %>%
  pairwise_t_test(TOC_m ~ Station, paired=FALSE, alternative= "less")
yp2 <- yp[!is.na(yp$sUVa_m),]
pwc2 <- yp %>%
  pairwise_t_test(sUVa_m ~ Station, paired=FALSE, alternative= "two.sided")
yp3 <- yp[!is.na(yp$THg_m),]
pwc3 <- yp3 %>%
  pairwise_t_test(THg_m ~ Station, paired=FALSE, alternative= "greater")
yp4 <- yp[!is.na(yp$THGTOC_m2),]
pwc4 <- yp4 %>%
  pairwise_t_test(THGTOC_m2 ~ Station, paired=FALSE, alternative= "greater")

pwc1
pwc2
pwc3
pwc4

 

pwc6 <- y16 %>%
  pairwise_t_test(sUVa_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc7 <- y16 %>%
  pairwise_t_test(E2_E3_m ~ Station, paired=FALSE, alternative= "two.sided")
pwc6
pwc7

```

```{r}
yp$Year=format(as.Date(yp$Date, format="%d.%m.%Y"),"%Y")

ggplot(subset(yp, Station=="Inlet"))+
  #geom_point(aes(x=Date, y=Disch), size=5)+
  geom_line(aes(x=Date, y=Disch, colour=Station))+
  #geom_point(aes(x=Date, y=TOC_m), colour="black")+
  #geom_smooth(aes(x=Date, y=Disch, colour=Station), shape=16, size=1)+
  #geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year, scales="free")+
  scale_y_log10()


ggplot(subset(yp, Year %in% c(2016, 2017, 2018, 2019, 2020, 2021)))+
  geom_line(aes(x=Date, y=Disch, shape=Station, colour=Station), size=1)+
  geom_point(aes(x=Date, y=Disch, shape=Station, colour=Station), size=2)+
  #geom_line(aes(x=month_day, y=TOC_m, colour=Station))+
  #geom_smooth(aes(x=Date, y=Disch, colour=Station), shape=16, size=1)+
  #geom_point(aes(x=month_day, y=DOC_s, colour=Station), shape=17, size=5)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  theme_bw()+
  labs(y = "TOC (mg/L)")+
   theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    axis.text.x = element_text(size=12, angle = 45, hjust = 1),
    axis.text.y = element_text(size=12),
    strip.text = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0))
  )+
  facet_grid(.~ Year, scales="free")
```

Looking at the data without decomposing
```{r}
ggplot(yp)+
  geom_line(aes(x=Date, y=sUVa_m), colour="red")+
  geom_line(aes(x=Date, y=THg_m))+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Timeline, seasonality and correlations
- first, regularize the data (do we need to different since less Hg data?)
- seasonal decomposition
- correlation analysis
- keep in mind possible lag in time
- Hg shorter time frame than TOC
```{r Decompose}
dx_ts_inlet <- subset(yp, Station == "Inlet") #daily?
dx_ts_outlet <- subset(yp, Station == "Outlet") #daily?

# 1. Create a new data.frame that has a full sequence of dates, one for inlet and one for outlet
full_out <- data.frame(
  Date = seq(min(dx_ts_outlet$Date), max(dx_ts_outlet$Date), by = "day"))
full_in <- data.frame(
  Date = seq(min(dx_ts_inlet$Date), max(dx_ts_inlet$Date), by = "day"))

# 2. Merge with the old one on the `date` column
new_ts_out <- merge(full_out, dx_ts_outlet, by = "Date", all.x = TRUE)
new_ts_in <- merge(full_in, dx_ts_inlet, by = "Date", all.x = TRUE)

# 3. Some values are now missing - replace them with 0
#library(zoo)
#For TOC and sUVa 
new_ts_out1 <- new_ts_out[c(1:2189),] #Specify dates to be included to avoid leading or ending NAs
new_ts_in1 <- new_ts_in[c(1:2165),]    #Specify dates to be included to avoid leading or ending NAs

#For THg
new_ts_out2 <- new_ts_out[c(30:2189),] #Specify dates to be included to avoid leading or ending NAs
new_ts_in2 <- new_ts_in[c(1:2165),]    #Specify dates to be included to avoid leading or ending NAs

#For discharge
new_ts_out3 <- new_ts_out[c(1:2189),] #Specify dates to be included to avoid leading or ending NAs
new_ts_in3 <- new_ts_in[c(1:2165),]    #Specify dates to be included to avoid leading or ending NAs

#Hg-to-DOC
new_ts_out4 <- new_ts_out[c(30:2189),] #Specify dates to be included to avoid leading or ending NAs
new_ts_in4 <- new_ts_in[c(1:2165),]    #Specify dates to be included to avoid leading or ending NAs

new_ts_out1$inter_TOC_out <- na.approx(new_ts_out1$TOC_m)
new_ts_in1$inter_TOC_in <- na.approx(new_ts_in1$TOC_m)

new_ts_out2$inter_THg_out <- na.approx(new_ts_out2$THg_m, na.rm = TRUE)
new_ts_in2$inter_THg_in <- na.approx(new_ts_in2$THg_m)

new_ts_out1$inter_sUVa_out <- na.approx(new_ts_out1$sUVa_m, na.rm = TRUE)
new_ts_in1$inter_sUVa_in <- na.approx(new_ts_in1$sUVa_m)

new_ts_out3$inter_Dis_out <- na.approx(new_ts_out3$Disch, na.rm = TRUE)
new_ts_in3$inter_Dis_in <- na.approx(new_ts_in3$Disch)

new_ts_out4$inter_THgTOC_out <- na.approx(new_ts_out4$THGTOC_m2, na.rm = TRUE)
new_ts_in4$inter_THgTOC_in <- na.approx(new_ts_in4$THGTOC_m2)

#simple plots
qplot(x=Date, y=TOC_m,data=dx_ts)+
  facet_grid(Station~.)

qplot(x=Date, y=THg_m, data=dx_ts)+
  facet_grid(Station~.)

qplot(x=Date, y=sUVa_m, data=dx_ts)+
  facet_grid(Station~.)

qplot(x=Date, y=Disch, data=yp)+ #some high values that may help explain the TOC?
  facet_grid(Station~.)

#try decompose
#https://www.statology.org/how-to-perform-time-series-decomposition-r/
A1_out <- ts(new_ts_out1, frequency=365) #TOC and sUVa data outlet
A1_inl <- ts(new_ts_in1, frequency=365)  #TOC and sUVa data inlet

B1_out <- ts(new_ts_out2, frequency=365) #Hg data outlet
B1_inl <- ts(new_ts_in2, frequency=365)  #Hg data inlet

C1_out <- ts(new_ts_out3, frequency=365) #Discharge data outlet
C1_inl <- ts(new_ts_in3, frequency=365)  #Discharge data inlet

D1_out <- ts(new_ts_out4, frequency=365) #Discharge data outlet
D1_inl <- ts(new_ts_in4, frequency=365)  #Discharge data inlet

library(stats)
#decomp <- stl(A1_out[,5], s.window = "periodic") #date is not on x axis
res1<- (stl(ts(A1_out[,9],frequency=365,start=c(2016,2 )),s.window="periodic",robust=TRUE) )#fixes date. TOC outlet
res2<- (stl(ts(A1_inl[,9],frequency=365,start=c(2016,31)),s.window="per",robust=TRUE) )#fixes date. TOC inlet

res3<- (stl(ts(B1_out[,9],frequency=365,start=c(2016,31 )),s.window="per",robust=TRUE) )#fixes date. Hg outlet
res4<- (stl(ts(B1_inl[,9],frequency=365,start=c(2016,31)),s.window="per",robust=TRUE) )#fixes date. Hg inlet

res5<- (stl(ts(A1_out[,10],frequency=365,start=c(2016,2 )),s.window="per",robust=TRUE) )#fixes date. sUVa outlet
res6<- (stl(ts(A1_inl[,10],frequency=365,start=c(2016,31)),s.window="per",robust=TRUE) )#fixes date. sUVa inlet

res7<- (stl(ts(C1_out[,9],frequency=365,start=c(2016,2 )),s.window="periodic",robust=TRUE) )#fixes date. disch outlet
res8<- (stl(ts(C1_inl[,9],frequency=365,start=c(2016,31)),s.window="per",robust=TRUE) )#fixes date. disch inlet

res9<- (stl(ts(D1_out[,9],frequency=365,start=c(2016,2 )),s.window="periodic",robust=TRUE) )#fixes date. THg:DOC outlet
res10<- (stl(ts(D1_inl[,9],frequency=365,start=c(2016,31)),s.window="per",robust=TRUE) )#fixes date. THg:DOC inlet

plot(res1)+title("Langtjern TOC outlet")
plot(res2)+title("Langtjern TOC inlet")
plot(res3)+title("Langtjern Hg outlet")
plot(res4)+title("Langtjern Hg inlet")
plot(res5)+title("Langtjern sUVa outlet")
plot(res6)+title("Langtjern sUVa inlet")
plot(res7)+title("Langtjern water discharge outlet")
plot(res8)+title("Langtjern water discharge inlet")

plot(res9)+title("Langtjern Hg:DOC outlet")
plot(res10)+title("Langtjern Hg:DOC inlet")

df <- data.frame(matrix(unlist(res1), nrow=length(res1), byrow=TRUE))

#EXTRACT AND LOOK FOR CORRELATIONS
TOC_out <- as.data.frame(res1$time.series)
TOC_in <- as.data.frame(res2$time.series)
Hg_out <- as.data.frame(res3$time.series)
Hg_in <- as.data.frame(res4$time.series)
sUVa_out <- as.data.frame(res5$time.series)
sUVa_in <- as.data.frame(res6$time.series)
dis_out <- as.data.frame(res7$time.series)
dis_in <- as.data.frame(res8$time.series)

THgDOC_out <- as.data.frame(res9$time.series)
THgDOC_in <- as.data.frame(res10$time.series)

#ADD DATE COLUMN
# 2. Define the start date, number of periods, and frequency
start_date1 <- as.Date("2016-01-02")
start_date2 <- as.Date("2016-01-31")

num_periods1 <- nrow(TOC_out) # Match the number of rows in the dataframe. TOC and sUVA the dame
num_periods2 <- nrow(TOC_in) # Match the number of rows in the dataframe. TOC and sUVa the same
num_periods3 <- nrow(Hg_out) # Match the number of rows in the dataframe. Hg
num_periods4 <- nrow(Hg_in) # Match the number of rows in the dataframe. Hg
num_periods5 <- nrow(sUVa_out) # Match the number of rows in the dataframe. sUVa
num_periods6 <- nrow(sUVa_in) # Match the number of rows in the dataframe. sUVa
num_periods7 <- nrow(dis_out) # Match the number of rows in the dataframe. discharge
num_periods8 <- nrow(dis_in) # Match the number of rows in the dataframe. discharge

num_periods9 <- nrow(THgDOC_out) # Match the number of rows in the dataframe. discharge
num_periods10 <- nrow(THgDOC_in) # Match the number of rows in the dataframe. discharge

frequency <- "days" # Can be "days", "weeks", "months", "years", etc.

# 3. Generate the date sequence using seq()
date_sequence1 <- seq(from = start_date1, by = frequency, length.out = num_periods1)
date_sequence2 <- seq(from = start_date2, by = frequency, length.out = num_periods2)
date_sequence3 <- seq(from = start_date2, by = frequency, length.out = num_periods3)
date_sequence4 <- seq(from = start_date2, by = frequency, length.out = num_periods4)
date_sequence5 <- seq(from = start_date2, by = frequency, length.out = num_periods5)
date_sequence6 <- seq(from = start_date2, by = frequency, length.out = num_periods6)
date_sequence7 <- seq(from = start_date2, by = frequency, length.out = num_periods7)
date_sequence8 <- seq(from = start_date2, by = frequency, length.out = num_periods8)

date_sequence9 <- seq(from = start_date2, by = frequency, length.out = num_periods9)
date_sequence10 <- seq(from = start_date2, by = frequency, length.out = num_periods10)

# 4. Add the date sequence as a new column to the dataframe
TOC_out$Date <- date_sequence1
TOC_in$Date <- date_sequence2
Hg_out$Date <- date_sequence3
Hg_in$Date <- date_sequence4
sUVa_out$Date <- date_sequence5
sUVa_in$Date <- date_sequence6
dis_out$Date <- date_sequence7
dis_in$Date <- date_sequence8

THgDOC_out$Date <- date_sequence9
THgDOC_in$Date <- date_sequence10

# Align on overlapping dates
outlet <- merge(TOC_out[, c("Date", "seasonal", "trend", "remainder")],
            Hg_out[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_TOC", "_Hg"))

outlet2 <- merge(outlet[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC", "seasonal_Hg", "trend_Hg", "remainder_Hg")],
            sUVa_out[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_sUVa"))

# Rename specific columns
outlet2x <- outlet2 %>%
  rename(seasonal_sUVa =seasonal,
         trend_sUVa =trend,
         remainder_sUVa =remainder 
  )

outlet3 <- merge(outlet2x[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC", "seasonal_Hg", "trend_Hg", "remainder_Hg", "seasonal_sUVa", "trend_sUVa", "remainder_sUVa")],
            dis_out[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("x", "_dis"))

# Rename specific columns
outlet3x <- outlet3 %>%
  rename(seasonal_dis =seasonal,
         trend_dis =trend,
         remainder_dis =remainder 
  )

#Add THg:DOC
outlet4 <- merge(outlet3x[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC", "seasonal_Hg", "trend_Hg", "remainder_Hg", "seasonal_sUVa", "trend_sUVa", "remainder_sUVa", "seasonal_dis", "trend_dis", "remainder_dis" )],
            THgDOC_out[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_THgDOC"))

#REPEAT THE SAME TILPASNINGER HERE AS FOR OUTLET
inlet <- merge(TOC_in[, c("Date", "seasonal", "trend", "remainder")],
            Hg_in[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_TOC", "_Hg"))


inlet2 <- merge(inlet[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC","seasonal_Hg", "trend_Hg", "remainder_Hg")],
            sUVa_in[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_TOC", "_sUVa"))

# Rename specific columns
inlet2x <- inlet2 %>%
  rename(seasonal_sUVa =seasonal,
         trend_sUVa =trend,
         remainder_sUVa =remainder 
  )

inlet3 <- merge(inlet2x[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC","seasonal_Hg", "trend_Hg", "remainder_Hg", "seasonal_sUVa", "trend_sUVa", "remainder_sUVa")],
            dis_in[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_TOC", "_sUVa"))

# Rename specific columns
inlet3x <- inlet3 %>%
  rename(seasonal_dis =seasonal,
         trend_dis =trend,
         remainder_dis =remainder 
  )

#Add THg:DOC
inlet4 <- merge(inlet3x[, c("Date", "seasonal_TOC", "trend_TOC", "remainder_TOC", "seasonal_Hg", "trend_Hg", "remainder_Hg", "seasonal_sUVa", "trend_sUVa", "remainder_sUVa", "seasonal_dis", "trend_dis", "remainder_dis" )],
            THgDOC_in[, c("Date", "seasonal", "trend", "remainder")],
            by = "Date", suffixes = c("_THgDOC"))

#SEASONAL - out
cor(outlet$seasonal_TOC, outlet$seasonal_Hg, use = "complete.obs")
plot(outlet$seasonal_TOC, outlet$seasonal_Hg, use = "complete.obs")

cor(inlet$seasonal_TOC, inlet$seasonal_Hg, use = "complete.obs")
plot(inlet$seasonal_TOC, inlet$seasonal_Hg, use = "complete.obs")

cor(inlet3$seasonal_TOC, inlet$seasonal_dis, use = "complete.obs")
plot(inlet3x$seasonal_TOC, inlet3x$seasonal_dis, use = "complete.obs")


#TREND
cor(outlet$trend_TOC, outlet$trend_Hg, use = "complete.obs")
plot(outlet$trend_TOC, outlet$trend_Hg, use = "complete.obs")

cor(inlet$trend_TOC, inlet$trend_Hg, use = "complete.obs")
plot(inlet$trend_TOC, inlet$trend_Hg, use = "complete.obs")

```


```{r Plot decomposed data}
#NEED YEAR
inlet4$Year=format(as.Date(inlet$Date, format="%d.%m.%Y"),"%Y")
outlet4$Year=format(as.Date(outlet$Date, format="%d.%m.%Y"),"%Y")

#SEASONAL
ggplot(subset(outlet4, Year == 2018))+
  geom_line(aes(x=Date, y=seasonal_dis, colour="seasonal_dis"), linetype = "dashed", linewidth=1.5)+
  geom_line(aes(x=Date, y=seasonal_Hg, colour="seasonal_Hg"), linewidth = 1.5)+
  geom_line(aes(x=Date, y=seasonal_sUVa, colour="seasonal_sUVa"), linewidth = 1.5)+
  geom_line(aes(x=Date, y=seasonal_TOC, colour="seasonal_TOC"), linewidth = 1.2)+
  #geom_line(aes(x=Date, y=seasonal, colour="seasonal"))+
  scale_x_date(
    date_breaks = "1 month", # Sets a major tick mark every month
    date_labels = "%b"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
   theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 16, colour="black"),
        axis.text.y = element_text(size= 17, colour="black"),
        axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), size=17, colour="black"),        
        axis.title.x = element_blank(),        
        legend.text = element_blank(),#size = 12, colour="black"
        plot.title = element_text(size = 20))+
  ggtitle("Outlet - seasonal")+
  ylab("Seasonal component (centered)")+
  scale_colour_manual("", 
                      labels = c("DOC","sUVa254",  "Hg", "Water discharge"),
                      breaks = c("seasonal_TOC","seasonal_sUVa", "seasonal_Hg", "seasonal_dis"),
                      values = c("brown","orange","red", "blue")) +
  scale_y_continuous(labels = label_number(accuracy = 0.1))
   bnscale_y_log10()
  
  

  #  scale_colour_discrete(name = "Dose", labels = c("Hg", "sUVa254", "TOC", "Water discharge"))

  
ggplot(outlet4)+
  geom_line(aes(x=Date, y=seasonal_Hg), colour="red")+
  geom_line(aes(x=Date, y=seasonal_sUVa), colour="orange")+
  geom_line(aes(x=Date, y=seasonal), colour="purple")+
  #geom_line(aes(x=Date, y=seasonal_TOC), colour="darkgreen")+
  #geom_line(aes(x=Date, y=seasonal_dis), colour="blue")+
  scale_x_date(
    date_breaks = "2 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Outlet - seasonal")+
  theme(axis.title = element_blank())
  scale_y_log10()

#TREND
ggplot(inlet3x)+
  geom_line(aes(x=Date, y=trend_Hg), colour="red")+
  geom_line(aes(x=Date, y=trend_sUVa), colour="orange")+
  geom_line(aes(x=Date, y=trend_TOC), colour="darkgreen")+
  geom_line(aes(x=Date, y=trend_dis), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Inlet - trend (log10)")+
  theme(axis.title = element_blank())+
  scale_y_log10()

ggplot(outlet3x)+
  geom_line(aes(x=Date, y=trend_Hg), colour="red")+
  geom_line(aes(x=Date, y=trend_sUVa), colour="orange")+
  geom_line(aes(x=Date, y=trend_TOC), colour="darkgreen")+
  geom_line(aes(x=Date, y=trend_dis), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Outlet - trend (log10)")+
  theme(axis.title = element_blank())+
  scale_y_log10()

#DATA
ggplot(inlet3x)+
  geom_line(aes(x=Date, y=remainder_Hg), colour="red")+
  geom_line(aes(x=Date, y=remainder_sUVa), colour="orange")+
  geom_line(aes(x=Date, y=remainder_TOC), colour="darkgreen")+
  geom_line(aes(x=Date, y=remainder_dis), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Inlet - remainder")+
  theme(axis.title = element_blank())+
  scale_y_log10()

ggplot(outlet3x)+
  geom_line(aes(x=Date, y=remainder_Hg), colour="red")+
  geom_line(aes(x=Date, y=remainder_sUVa), colour="orange")+
  geom_line(aes(x=Date, y=remainder_TOC), colour="darkgreen")+
  geom_line(aes(x=Date, y=remainder_dis), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Outlet - remainder")+
  theme(axis.title = element_blank())+
  scale_y_log10()






#residual1  
ggplot(outlet3x)+
  geom_line(aes(x=Date, y=remainder_TOC), colour="red")+
  geom_line(aes(x=Date, y=remainder_dis))

ggplot(inlet)+
  geom_line(aes(x=Date, y=remainder_TOC), colour="red")+
  geom_line(aes(x=Date, y=remainder_Hg))

#seasonal
ggplot(outlet2)+
  geom_line(aes(x=Date, y=seasonal_TOC), colour="red")+
  geom_line(aes(x=Date, y=seasonal))

ggplot(inlet2)+
  geom_line(aes(x=Date, y=seasonal_TOC), colour="red")+
  geom_line(aes(x=Date, y=seasonal))

#trend
ggplot(outlet2)+
  geom_line(aes(x=Date, y=trend_TOC), colour="red")+
  geom_line(aes(x=Date, y=trend))

ggplot(inlet2)+
  geom_line(aes(x=Date, y=trend_TOC), colour="red")+
  geom_line(aes(x=Date, y=trend))

#residual  
ggplot(outlet2)+
  geom_line(aes(x=Date, y=remainder_TOC), colour="red")+
  geom_line(aes(x=Date, y=remainder))

ggplot(inlet2)+
  geom_line(aes(x=Date, y=remainder_TOC), colour="red")+
  geom_line(aes(x=Date, y=remainder))



#seasonal
ggplot(outlet2)+
  geom_line(aes(x=Date, y=seasonal_Hg))+
  geom_line(aes(x=Date, y=seasonal), colour="red")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Outlet: sUVa (red)~Hg (black)")

ggplot(inlet2)+
  geom_line(aes(x=Date, y=seasonal_Hg))+
  geom_line(aes(x=Date, y=seasonal), colour="red")+
  scale_x_date(
    date_breaks = "2 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Inlet: sUVa (red)~Hg (black)")


#MERGE inlet and outlet to have in the same figure
Langtjern <- merge(inlet4, outlet4, by="Date")

#Seasonal: inlet and outlet together
ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_TOC.x))+
  geom_line(aes(x=Date, y=seasonal_TOC.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("TOC: inlet (black)~ outlet (blue)")

ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_dis.x))+
  geom_line(aes(x=Date, y=seasonal_dis.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Dis: inlet (black)~ outlet (blue)")

ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_Hg.x))+
  geom_line(aes(x=Date, y=seasonal_Hg.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Hg: inlet (black)~ outlet (blue)")

ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_sUVa.x))+
  geom_line(aes(x=Date, y=seasonal_sUVa.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("sUVa: inlet (black)~ outlet (blue)")


ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_Hg.x))+
  geom_line(aes(x=Date, y=seasonal_Hg.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Hg: inlet (black)~ outlet (blue)")

ggplot(Langtjern)+
  geom_line(aes(x=Date, y=seasonal_dis.x))+
  geom_line(aes(x=Date, y=seasonal_dis.y), colour="blue")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Discharge: inlet (black)~ outlet (blue)")





#seasonalXXXXX
ggplot(outlet2)+
  geom_line(aes(x=Date, y=seasonal_TOC), colour="blue")+
  geom_line(aes(x=Date, y=seasonal), colour="green")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Outlet: sUVa (green)~TOC (blue)")

ggplot(inlet2)+
  geom_line(aes(x=Date, y=seasonal_TOC), colour="blue")+
  geom_line(aes(x=Date, y=seasonal), colour="green")+
  scale_x_date(
    date_breaks = "4 month", # Sets a major tick mark every month
    date_labels = "%b %Y"    # Formats the labels as 'Jan 2023', 'Feb 2023', etc.
  )+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
   ggtitle("Inlet: sUVa (red)~Hg (black)")



#trend
ggplot(outlet2)+
  geom_line(aes(x=Date, y=trend_Hg), colour="red")+
  geom_line(aes(x=Date, y=trend))

ggplot(inlet2)+
  geom_line(aes(x=Date, y=trend_Hg), colour="red")+
  geom_line(aes(x=Date, y=trend))

#residual  
ggplot(outlet2)+
  geom_line(aes(x=Date, y=remainder_Hg), colour="red")+
  geom_line(aes(x=Date, y=remainder))

ggplot(inlet2)+
  geom_line(aes(x=Date, y=remainder_Hg), colour="red")+
  geom_line(aes(x=Date, y=remainder))



library("ggpubr")
head(dx)
ggscatter(outlet2, y = "seasonal_Hg", x = "seasonal_TOC", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="outlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(inlet2, y = "seasonal_Hg", x = "seasonal_TOC", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="inlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(outlet2, y = "seasonal_Hg", x = "seasonal", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="outlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(inlet2, y = "seasonal_Hg", x = "seasonal", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="inlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")
  facet_grid(.~Station)


  
  
ggscatter(outlet2, y = "trend_Hg", x = "trend_TOC", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="outlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(inlet2, y = "trend_Hg", x = "trend_TOC", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="inlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(outlet2, y = "trend_Hg", x = "trend", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="outlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")

  ggscatter(inlet2, y = "trend_Hg", x = "trend", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson", title="inlet")
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")
  facet_grid(.~Station)
  
  
  
  
  
  
  
  
  
  



cor(inlet$seasonal_TOC, outlet$seasonal_TOC, use = "complete.obs", by="Date")


cor(outlet$seasonal_TOC, inlet$seasonal_TOC) 
nrow(outlet)


#extract to assess correlation
TOC_o_s1 <- decomposed_TOC_out$seasonal;  TOC_o_t1 <- decomposed_TOC_out$trend;  TOC_o_r1 <- decomposed_TOC_out$random
TOC_i_s2 <- decomposed_TOC_in$seasonal;  TOC_o_t2 <- decomposed_TOC_in$trend;  TOC_o_r2 <- decomposed_TOC_in$random

Hg_o_s1 <- decomposed_Hg_out$seasonal;  Hg_o_t1 <- decomposed_Hg_out$trend;  Hg_o_r1 <- decomposed_Hg_out$random
Hg_i_s2 <- decomposed_Hg_in$seasonal;  Hg_i_t2 <- decomposed_Hg_in$trend;  Hg_i_r2 <- decomposed_Hg_in$random

#Compensate for shorter Hg
n <- min(length(Hg_o_s1), length(TOC_o_s1))
ts1_cut <- Hg_o_s1[1:n]
ts2_cut <- TOC_o_s1[1:n]

cor(ts1_cut, ts2_cut, use = "complete.obs")
plot(ts1_cut, ts2_cut)

df_TOC_out <- as.data.frame(decomposed_TOC_out[c("x", "seasonal", "trend", "random")])
df_TOC_in <- as.data.frame(decomposed_TOC_in[c("x", "seasonal", "trend", "random")])
df_THg_out <- as.data.frame(decomposed_Hg_out[c("x", "seasonal", "trend", "random")])
df_THg_in <- as.data.frame(decomposed_Hg_in[c("x", "seasonal", "trend", "random")])

ggplot(df_TOC_out)



library(dplyr)
TOC_o_s1x <- as.numeric(TOC_o_s1)
Hg_o_s1x <- as.numeric(Hg_o_s1)
joined_data <- inner_join(TOC_o_s1x, Hg_o_s1x, by = "Date")

print(joined_data)

cor.test(as.numeric(ts1_shortened), as.numeric(ts2_shortened), method = "pearson")
plot(as.numeric(TOC_o_s1), as.numeric(Hg_o_s1))


#contniue here!!! get date back to data and look at correlation

#correlate trends
#https://www.datascienceinstitute.net/ai-data-science-blog/time-series-decomposition-in-r





#simple plots
qplot(x=Date, y=TOC_m, data=dx_ts)+
  facet_grid(Station~.)
qplot(x=Date, y=THg_m, data=dx_ts)+
  facet_grid(Station~.)

#convert to ts time series format which is basic R
library(stats) 
# Unleash the time series magic!
ts_sales <- ts(dx_ts$TOC_m, frequency = 365)  # Daily data

# Admire your creation:
print(ts_sales)


plot(TOC_m_df, type="l")
plot(dx$TOC, type="l")


#TIMELINE VISUALISATION: https://www.r-bloggers.com/2023/11/time-series-analysis-in-r-how-to-read-and-understand-time-series-data/
typeof(dx$THg_m)
#TOC )2016 - 2026
ggplot(dx, aes(x = Date, y = TOC_m)) +
  geom_line()+
  facet_grid(Station~.)

#Hg (2016-2022)
ggplot(filter(dx, !is.na(THg_m)), aes(x = Date, y = THg_m)) +
  geom_line()+
  facet_grid(Station~.)

library(zoo)


# Create a numeric feature
dx$month_num <- as.numeric(dx$month)

# Fit a linear regression model
model <- lm(TOC_m ~ Month_num, data = dx)
# Get predictions
data$Trend_Line <- predict(model, newdata = data)

head(data[c("Month", "Passengers", "Trend_Line")], 12)
    
#https://a-little-book-of-r-for-time-series.readthedocs.io/en/latest/src/timeseries.#html

plot.ts(dx$TOC_m)

#We can see from the time plot that this time series could probably be described #using an additive model, since the random fluctuations in the data are roughly #constant in size over time.
TOC_m_df <- dx[,c(4:5)]
plot.ts(TOC_m_df)

TOC_m_df2 <- na.omit(TOC_m_df)

TOC_m_ts <- ts(TOC_m_df, frequency=12)
TOC_m_df
birthstimeseriescomponents <- decompose(TOC_m_ts)

```

## Sumamry table
```{r create summary table}
library(dplyr)
library(rbind)

dxy <- do.call(rbind.data.frame, dx)

dxz <- subset(dx, Year %in% c(2016, 2017, 2018, 2019, 2020, 2021))

#dx_sum <- dx[c("Station", "TOC_m", "THg_m", "sUVa_m", "THg_m")]
dx_sum <- select(dxz, "Station", "TOC_m", "THg_m", "sUVa_m", "THg_m", "THGTOC_m2", "Year", "month")

typeof(dx_sum)
class(dx_sum$TOC_m)

inl <- subset(dx_sum, Station  == "Inlet")
ou <- subset(dx_sum, Station  == "Outlet")
#length(inl$THgTOC_m, na.omit=TRUE)

colSums(!is.na(ou))

nrow(inl)

inl %>% count(THg_m)

sum <- dx_sum %>% 
  group_by(Station, month) %>% 
  summarise(across(.cols  = where(is.numeric), .f = list(mean = mean, min = min, max = max, sd = sd), na.rm = TRUE))
  
#sum_n <- dx_sum %>% 
#  group_by(Station) %>% 
#  length(across(.cols  = where(is.numeric)))

sum_n <- dx_sum %>% 
group_by(Station, month) %>%
  summarise(
    across(
      where(is.numeric),
      ~ sum(!is.na(.x)),
      .names = "n_{.col}"
    ),
    .groups = "drop"
  )

write.csv(sum, "summary monitoing.csv")
```

Summary tables for intra and inter-annual variability
- SD
- CV = SD/x *100

inl <- subset(dx_sum, Station  == "Inlet")
ou <- subset(dx_sum, Station  == "Outlet")
```{r}
#intra - for year year, what is the variability?
sumY <- dx_sum %>% 
  group_by(Station, Year) %>% 
  summarise(across(.cols  = where(is.numeric), .f = list(mean = mean, sd = sd), na.rm = TRUE))
#inter - for year month, what is the variabilitY?

sumM <- dx_sum %>% 
  group_by(Station, month) %>% 
  summarise(across(.cols  = where(is.numeric), .f = list(mean = mean, sd = sd), na.rm = TRUE))

write.csv(sumY, "Year_variability.csv")
write.csv(sumM, "Month_variability.csv")
```


#GAMM modelling
```{r transformations}
Gdx_inl <-yp %>% filter(Station == "Inlet") %>%
  arrange(Station, Date) %>%
  mutate(
    doy = yday(Date),
    y = year(Date),
    t = as.numeric(difftime(Date, min(Date), units = "days")),
    lnHg = log(THg_m),               # often stabilizes variance; check first
    lnTOC = log(TOC_m),          # log1p if zeros are possible
    sUVa_sc = scale(sUVa_m),       # standardize for interpretability
    Q_sc = scale(Disch),             # if available
    site = factor(Station)
  )

Gdx_out <-yp %>% filter(Station == "Outlet") %>%
  arrange(Station, Date) %>%
  mutate(
    doy = yday(Date),
    y = year(Date),
    t = as.numeric(difftime(Date, min(Date), units = "days")),
    lnHg = log(THg_m),               # often stabilizes variance; check first
    lnTOC = log(TOC_m),          # log1p if zeros are possible
    sUVa_sc = scale(sUVa_m),       # standardize for interpretability
    Q_sc = scale(Disch),             # if available
    site = factor(Station)
  )

#library(mgcv)
```

##GAMM - WITH DISCHARGE MANUSCRIPT
SIMPLE MODEL
```{r}
#library(nlme)
m1_gamm_inl <- gamm(
  lnHg ~ s(lnTOC, k = 6, bs = "cs") +
         s(sUVa_sc, k = 6, bs = "cs") +
         s(Q_sc, k = 6, bs = "cs") +
         s(doy, bs = "cc", k = 24) +
         s(t, k = 12, bs = "cs"),
  data = Gdx_inl,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t)
  #correlation = corAR1(form = ~ t)
)

# 1) Smooths & significance
summary(m1_gamm_inl$gam)
plot(m1_gamm_inl$gam, pages = 1, shade = TRUE)

# 2)  Residuals vs Fitted
plot(m1_gamm_inl$lme)

# 3) Normality of Residuals
gam.check(m1_gamm_inl$gam)

# 4) Residual versus time
plot(resid(m1_gamm_inl$lme), type="l")

# 5 Autocorrelation Check (ACF of residuals). Residual correlation
acf(resid(m1_gamm_inl$lme, type = "normalized"), lag.max = 24)

# 6) Concurvity
concurvity(m1_gamm_inl$gam, full = TRUE)

# Summary of mixed model (AR1, random effects)
summary(m1_gamm_inl$lme)


#OUTLET
m1_gamm_out <- gamm(
        lnHg ~ s(lnTOC, k = 6, bs = "cs") +
         s(sUVa_sc, k = 6, bs = "cs") +
         s(Q_sc, k = 6, bs = "cs") +
         s(doy, bs = "cs", k = 24) +
         s(t, k = 12, bs = "cs"),
  data = Gdx_out,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t)
  #correlation = corAR1(form = ~ t)
)

# Smooths & significance
summary(m1_gamm_out$gam)
plot(m1_gamm_inl$gam, pages = 1, shade = TRUE)
plot(m1_gamm_inl$lme)
plot(m1_gamm_out$lme)

# Residual correlation
acf(resid(m1_gamm_out$lme, type = "normalized"), lag.max = 24)

#ACF of the normalized residuals from the GAMM showed no substantial autocorrelation; nearly all lags were within 95% bounds and no systematic decay was evident, indicating that the specified correlation structure adequately accounted for temporal dependence. 

# Basis checks (interpret cautiously for `gamm`)
gam.check(m1_gamm_out$gam)

# Concurvity
concurvity(m1_gamm_inl$gam, full = TRUE)

# Summary of mixed model (AR1, random effects)
summary(m1_gamm_inl$lme)


# Nicer plots
#library(gratia)
draw(m1_gamm_inl$gam)
draw(m1_gamm_out$gam)

```

Seasonality
-  If you want shrinkage/automatic removal of weak smooths, use a shrinkage basis (bs="cs"
```{r}
Gdx3_inl <- Gdx_inl %>%
  mutate(season = case_when(
    doy %in%  90:151 ~ "spring",     # adjust to local hydrology (snowmelt)
    doy %in% 152:243 ~ "summer",
    doy %in% 245:334 ~ "autumn",
    TRUE             ~ "winter"
  )) %>%
  mutate(season = factor(season, levels = c("winter","spring","summer","autumn")))

Gdx3_out <- Gdx_out %>%
  mutate(season = case_when(
    doy %in%  90:151 ~ "spring",     # adjust to local hydrology (snowmelt)
    doy %in% 152:243 ~ "summer",
    doy %in% 245:334 ~ "autumn",
    TRUE             ~ "winter"
  )) %>%
  mutate(season = factor(season, levels = c("winter","spring","summer","autumn")))

#library(mgcv)
#library(nlme)

m2_gamm_inl <- gamm(
  lnHg ~ season +                                  # gives each season its Hg-intercepts
         s(lnTOC,  by = season, k = 6, bs = "cs") +# shrinkage basis instead of select=TRUE
         s(sUVa_sc,by = season, k = 6, bs = "cs") + #cs= shrinkage cubic regression spline basis.
         s(Q_sc,   by = season, k = 6, bs = "cs") +  
         s(doy, bs = "cc", k = 20)+ #k = 24 is common for seasonal signal; Cyclic cubic spline.
         s(t, k = 10, bs = "cs"), #why cs here? It helps for autocorrelation. But if you are certain a trend exists and don’t want it to vanish, you can switch to bs="cr" (no shrinkage), but keep REML.
  data = Gdx3_inl,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t)) # distance-based (continuous-time) AR(1)

# Smooths & significance
summary(m2_gamm_inl$gam)
gam.check(m2_gamm_inl$gam)

# Residual correlation
acf(resid(m2_gamm_inl$lme, type = "normalized"), lag.max = 24)
# Concurvity
concurvity(m1_gamm_inl$gam, full = TRUE)


m2_gamm_out <- gamm(
  lnHg ~ season +                                  # gives each season its Hg-intercepts
         s(lnTOC,  by = season, k = 6, bs = "cs") +# shrinkage basis instead of select=TRUE
         s(sUVa_sc,by = season, k = 6, bs = "cs")+ #cs= shrinkage cubic regression spline basis.
         s(Q_sc,   by = season, k = 6, bs = "cs")+
         s(doy, bs = "cc", k = 20)+ #k = 24 is common for seasonal signal; Cyclic cubic spline.
         s(t, k = 8, bs = "cs"), #why cs here? It helps for autocorrelation. But if you are certain a trend exists and don’t want it to vanish, you can switch to bs="cr" (no shrinkage), but keep REML.
  data = Gdx3_out,
  method = "REML",
  correlation = nlme::corCAR1(form = ~ t))  # distance-based (continuous-time) AR(1)
  #correlation = corAR1(form = ~ t)  # AR(1) over time index t. Accounts for lag‑1 residual autocorrelation typical in weekly–monthly series

# Smooths & significance
summary(m2_gamm_out$gam)
gam.check(m2_gamm_out$gam)

# Residual correlation
acf(resid(m2_gamm_out$lme, type = "normalized"), lag.max = 24)
# Concurvity
concurvity(m1_gamm_out$gam, full = TRUE)



1# Nicer plots
#library(gratia)
draw(m2_gamm_inl$gam)
draw(m2_gamm_out$gam)


```





  
#geom_point(aes(x=Date, y=sUVa_s, colour=Station), size=5, shape=17)+
geom_smooth(aes(x=Date, y=sUVa_s, colour=Station), size=2)
#Correlation plots
library("ggpubr")
head(dx)
ggscatter(dx, x = "sUVa_m", y = "sVISa_m", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab ="Molecular SIze (sUVa)" , ylab ="THg:DOC (ng/mg)")
  facet_grid(.~Station)
    facet_grid(.~year(dx$Date), scales="free_x")
    
head(dx)
    #IMPORTANT!!! Correlation plot for Hg and DOM 
    ggplot(inlet, aes(labile_m, THGTOC_m))+
      geom_point(size=6)+
      geom_smooth(method = "lm", se = FALSE, colour="black")
      facet_grid(. ~ factor(Station), scales="free")
      xlim(4, 6)
      
      inlet <- subset(dx, Station == "Inlet")
      outlet <- subset(dx, Station == "Outlet")
    head(dx)
      #Correlation analysis
      library(dplyr)
      b <- select(dx, TOC_m, THg_m, MeHg_m, sUVa_m, sVISa_m, SAR_m, THg_m, labile_m, THGTOC_m, MeHGTOC_m)
      library(Hmisc)
      rcorr(as.matrix(b), type="pearson")
      class(d)
      library("PerformanceAnalytics")
      chart.Correlation(b, histogram=TRUE, pch=19)
      head(d)
      
      
      #MAKKING THE CORRELATION ASSESSMENT
      library(tidyverse)
      library(ggpubr)
      theme_set(theme_pubr())
      ggplot(outlet, aes(x = SAR_m, y = THGTOC_m)) +
        geom_point() +
        stat_smooth()+
        stat_smooth(method = lm)
      
model <- lm(THGTOC_m ~ sVISa_m, data = dx)
model  
summary(model)



#STATISTICAL TESTING
res.aov <- aov(THGTOC_m ~ Station, data = dx)
# Summary of the analysis
summary(res.aov)
TukeyHSD(res.aov)







#Correlation with SENSOR LAKE DATA
dy <- read.table("Langtjern_weather.txt", sep = "\t", header = TRUE) #load data
dy$Date <- as.Date(dy$SampleDate,format = "%d.%m.%Y") #convert to date column
df <- dy[ -c(1:5) ] #removed uneccesary columns
head(df)

#to make monthly summaries
dff = df %>%
  mutate(month = format(Date, "%m"), year = format(Date, "%Y")) %>%
  group_by(month, year) %>%
  summarise(sun = mean(Straaling),
            airT = mean(AirTemp),
            Prec = mean(Precip),
            Hum = mean(Humidity))

plot(dff)
head(dff)
#Aggreger per dag
library(dplyr)
(out <- df %>%
    group_by(Date) %>%
    summarize(Sun=sum(Straaling), 
              Precip=sum(Precip),
              Precip_ave=mean(Precip),
              AirTemp=mean(AirTemp),
              Humidity = mean(Humidity)))

head(df)

#PLOTTING with time on x-axis and year-facets
ggplot(out)+
  geom_point(aes(x=Date, y=Precip), shape=16, size=5)
  geom_smooth(aes(x=Date, y=Precip), shape=16, size=2)+
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  facet_grid(.~year(dx$Date), scales="free_x")+
  theme_bw()
ylim(2.5, 6)
scale_y_continuous(limit=c(3, 4))
geom_point(aes(x=Date, y=labile_s, colour=Station), shape=17, size=5)+
  head(dx)






ou <- head(out)
dfff <- na.omit(dff)
head(dfff)
ou = dfff
nrow(sun)
head(ou)
#Make indivudual datafiles as ts with correct frequency
sun <- ts(ou$sun, frequency=12, start=2016, end=2019)
pre <- ts(ou$Prec, frequency=12, start=2016, end=2019)
airT <- ts(ou$airT, frequency=12, start=2016, end=2019)
Humi <- ts(ou$Hum, frequency=12, start=2016, end=2019)

plot(sunny)
sunny <- decompose(sun)
precip <- decompose(pre)
AirTemp <- decompose(airT)
Humidity <- decompose(Humi)
plot(Humidity)
plot(AirTemp)
plot(precip)
plot(sunny)
ts.plot(pre)

#then YOU CAN SUBTRACT E.G. THE SEAOSNAL COMPONENT FROM THE DATA
birthsSeasonAdj <- sun - sunny$seasonal- sunny$random
plot.ts(birthsSeasonAdj)

#Make correspodning individual ts with correct frequency
#here we have inlet and outlet. can look at these individually and at the difference bwteen the two
newdata <- subset(dx,  Station =="Outlet")
sUVa <- na.omit(newdata$sUVa_m)
sUVaTS <- ts(sUVa, frequency=365, start=2016, end=2019)

decomsUVa <- decompose(sUVaTS)
plot(decomsUVa)

suva <- sUVaTS - decomsUVa$seasonal - decomsUVa$random
plot.ts(suva)
plot(suva~birthsSeasonAdj)

abline(lm(suva~birthsSeasonAdj))

ab <- lm(suva~birthsSeasonAdj)
summary(ab)

#What about doing monthly aggregates???








#PLoT værdata med eventuelt langtjern data
ggplot()+
  geom_smooth(dx, mapping=aes(x=Date, y=Precip), shape=16, size=2)+
  geom_point(dx, mapping=aes(x=Date, y=TOC_m, colour=Station), shape=16, size=3)
  scale_x_date(labels = date_format("%b"),breaks = date_breaks("2 months"))+
  scale_y_continuous(name = "Radiation", 
                     sec.axis = sec_axis(~.*10, name = "TOC"))
  facet_grid(.~year, scales="free_x")+
  theme_bw()
  

linearMod <- lm(_m ~ TOC_m, data=dx)
print(linearMod)
summary(linearMod)
